---
title: "Casual Inference Notebook"
output:
  html_document:
    df_print: paged
---

In this assignment we will exercise causal inference on the RHC (Right Heart Catheterization) dataset This dataset was used in Connors et al. (1996): The effectiveness of RHC in the initial care of critically ill patients. J American Medical Association 276:889-897. The dataset pertains to day 1 of hospitalization, i.e., the "treatment" variable swang1 is whether or not a patient received a RHC (also called the Swan-Ganz catheter) on the first day in which the patient qualified for a specific research study called SUPPORT.

What should you submit? You need to submit:
1. This Notebook (in .Rmd format) after filling in all your answers inside this notebook. You are required to do an action (write or fill-in code or write documentation) at the places marked by an arrow (=>). This arrow appears within the comments (and it will be preceeded by "#") or outside the comments in the text (without "#").
2. The Notebook in html format with all the results.

Please provide information about your group:
==> Student names: 
==> Student numbers:

#TODO: names, student numbers and disclaimer that all em dashes used were human written haha?

# Closed part (70% of the grade)

## Understand the dataset:
You can download the dataset from the course canvas. It is called "rhc.csv". Please Open https://hbiostat.org/data/repo/rhc.html and take a good look at the variables. Make sure you inspect the variables included in the dataset (they appear in a tabel below the text on the website). You can refer to the table whenever needed. 

Packages: install packages if needed
# ```{r}
# install.packages("tableone")
# install.packages("Matching")
# # Recommended: add "dplyr" if you have used it before to make efficient preprocessing.
# ```

## Load packages
```{r}
library(tableone) # This library allows us to calculate summary statistics about the variables including the SMD (standardized mean difference).
library(Matching) # This library will do the matching
```

## Read database
```{r}
# Download rhc dataset from CANVAS and read it
rhc <- read.csv("rhc.csv")
```

## Understanding the dataset  
Tip: Look at the data, use the "View" command on the dataset "rhc"
```{r}
print("===================================================")
# => how many rows does rhc contain? write a print command
# => What are the variable names? write a print command
number_rows <- nrow(rhc)
column_names <- names(rhc)
sprintf("There are %d rows in the rhc dataframe.", number_rows)
sprintf("The following are the column names in the rhc dataframe:")
print(column_names)
print("===================================================")
```

For the assignment the TREATMENT variable is swang1 and the OUTCOME variable is death.

## Preprocess
If you look at the values of the dichotomous variables you will see that they are factors (in other words, categorical variables). We would like to work with numeric variables in this exercise so we need to perform some conversions.

Let's start with the treatment and outcome variables. Create an "outcome.died" and "swang.treatment" variables with the numeric values 0 and 1. Later we will deal with the rest.
```{r}
rhc$outcome.died <- as.numeric(rhc$death=='Yes')  # Again you might want, instead, to do all preprocessing with dplyr if you prefer
rhc$treatment.swang <- as.numeric(rhc$swang1=='RHC')
```

### Check

Let's take a quick look into the dataset using these variables
```{r}
print("===================================================")
# => What is the mean of outcome.died in the whole data set?
mean_outcome_died <- mean(rhc$outcome.died)
sprintf("Mean of outcome.died in whole data set is: %.4f", mean_outcome_died)

# => How many subjects are treated?
num_treated <- sum(rhc$treatment.swang)
sprintf("There are %d subjects treated.", num_treated)

# =>  What is the mean mortality (outcome.died) in this treated group?
mean_mortality_treated <- mean(rhc$outcome.died[rhc$treatment.swang == 1])
sprintf("Mean mortality in treated group is: %.4f", mean_mortality_treated)

# => How many subjects are not treated (i.e. in the control group)?
num_not_treated <- sum(rhc$treatment.swang == 0)
sprintf("There are %d subjects not treated.", num_not_treated)

# => What is the mean mortality in this control group?
mean_mortality_not_treated <- mean(rhc$outcome.died[rhc$treatment.swang == 0])
sprintf("Mean mortality in not treated group is: %.4f", mean_mortality_not_treated)

print("===================================================")
```


The variable cat1 is a factor (categorical variable) containing the primary disease category for the patient.
```{r}
print("===================================================")
# => How many different categories are in Cat1? Tip: use the levels() function on this variable.
num_levels_cat1 <- length(levels(rhc$cat1))
class <-class(rhc$cat1)
sprintf("Cat1 is actually a %s class.", class)
# Convert to factor:
print("Converting to factor first, then counting number of different categories:")
cat1_factor = as.factor(rhc$cat1)
print("The different cat1 categories:")
print(levels(cat1_factor))
num_levels_cat1 <- length(levels(cat1_factor))
sprintf("Cat1 has %d different categories.", num_levels_cat1)
print("===================================================")
```

We would like to "translate" Cat1 into a set of variables each with the numeric values of 0/1. For example for the value of "ARF" we will create a variable "ARF" that will be 1 if rhc$cat1=='ARF'.

```{r}
rhc$ARF <- as.numeric(rhc$cat1=='ARF')
print("===================================================")
# => Do the same (as we did for ARF) also for all other values of cat1 but ignore the category "Missing".
rhc$CHF             <- as.numeric(rhc$cat1 == 'CHF')
rhc$Cirrhosis       <- as.numeric(rhc$cat1 == 'Cirrhosis')
rhc$Colon.Cancer    <- as.numeric(rhc$cat1 == 'Colon Cancer')
rhc$Coma            <- as.numeric(rhc$cat1 == 'Coma')
rhc$COPD            <- as.numeric(rhc$cat1 == 'COPD')
rhc$Lung.Cancer     <- as.numeric(rhc$cat1 == 'Lung Cancer')
rhc$MOSF.Malignancy <- as.numeric(rhc$cat1 == 'MOSF w/Malignancy')
rhc$MOSF.Sepsis     <- as.numeric(rhc$cat1 == 'MOSF w/Sepsis')
 print("===================================================")
```

## Create data frame with variables of interest

==> Create a data frame and call it rhc.small with the following variables of interest:
All the underlying categories of "cat1" (ARF etc) that you created before,   "cardiohx", "chfhx", "dementhx", "psychhx",  "chrpulhx", "renalhx", "liverhx",  "gibledhx", "malighx",  "immunhx",  "transhx",  "amihx", "age", "sex", "edu", "das2d3pc", "aps1", "scoma1", "meanbp1", wblc1", "hrt1",     "resp1", "temp1", "pafi1", "alb1", "hema1", "bili1", "crea1", "sod1", "pot1",     "paco21", "ph1", "wtkilo1", "dnr1", "resp", "card", "neuro", "gastr",  "renal", "meta", "hema", "seps", "trauma",  "ortho", "adld3p", "urin1", "treatment.swang", "outcome.died".

```{r}
# rhc.small <- => Create this data frame
rhc.small <- rhc[ , c("ARF", "CHF", "Cirrhosis", "Colon.Cancer", "Coma", "COPD",
                      "Lung.Cancer", "MOSF.Malignancy", "MOSF.Sepsis",
                      "cardiohx", "chfhx", "dementhx", "psychhx", "chrpulhx", "renalhx",
                      "liverhx", "gibledhx", "malighx", "immunhx", "transhx", "amihx",
                      "age", "sex", "edu", "das2d3pc",
                      "aps1", "scoma1", "meanbp1", "wblc1", "hrt1", "resp1", "temp1", "pafi1",
                      "alb1", "hema1", "bili1", "crea1", "sod1", "pot1", "paco21", "ph1", "wtkilo1",
                      "dnr1", "resp", "card", "neuro", "gastr", "renal", "meta", "hema", "seps",
                      "trauma", "ortho", "adld3p", "urin1",
                      "treatment.swang", "outcome.died")]
nrows <- nrow(rhc.small)
ncols <- ncol(rhc.small)
sprintf("rhc.small has %d rows and %d columns.", nrows, ncols)
```

Make sure you take a look at the definition of variables (on the URL given above) with unclear names to you. This data frame includes now all candidate variables from which we will select later for matching. However, before we jump into variable selection, let us continue data type conversion.

Check classes of these variables by using the str command
```{r}
str(rhc.small)
```

If you have "labelled" variables like labelled "int" and labelled "num" it means that the dataset was imported from SPSS. In such a case, simply change such variables into numeric variables. If you do not see such variables then skip this step.

```{r}
#To our understanding this is not the case with us, thus we skip this step.

#rhc.small$age <- as.numeric(rhc.small$age) # only if in your version of the data "age" was a "labelled" variable
print("===================================================")
# => Change all labelled variables that include numbers (integer or numeric) into numeric.
# print("===================================================")
```

For each  categorical variable with 2 values (such as sex) change it into a numeric variable by selecting one of the values (such as "female") to be 1 and the other zero.

```{r}
rhc.small$female <- as.numeric(rhc.small$sex=='Female')
print("===================================================")
# => Change any factor with two values into a 0/1 variable by selecting a value (such as "female" above) to be 1 and the other value will be zero.
rhc.small$dnr1   <- as.numeric(rhc.small$dnr1   == 'Yes')   
rhc.small$resp   <- as.numeric(rhc.small$resp   == 'Yes')  
rhc.small$card   <- as.numeric(rhc.small$card   == 'Yes')
rhc.small$neuro  <- as.numeric(rhc.small$neuro  == 'Yes')
rhc.small$gastr  <- as.numeric(rhc.small$gastr  == 'Yes')
rhc.small$renal  <- as.numeric(rhc.small$renal  == 'Yes')
rhc.small$meta   <- as.numeric(rhc.small$meta   == 'Yes') 
rhc.small$hema   <- as.numeric(rhc.small$hema   == 'Yes')
rhc.small$seps   <- as.numeric(rhc.small$seps   == 'Yes')
rhc.small$trauma <- as.numeric(rhc.small$trauma == 'Yes')
rhc.small$ortho  <- as.numeric(rhc.small$ortho  == 'Yes')

rhc.small$sex <- NULL #removing sex since it wasnt replaced in-place, without this it messed up the matching later on

#quick sanity check for whether all of the above worked
print(names(rhc.small)[sapply(rhc.small, is.factor)])
print("===================================================")
```
Before we start the analysis we need to be aware of the following issue: If you encounter a variable with missing values then you might run into a problem with the Match() function later on. For this exercise you can follow this advice:

1) If the proportion of missing is above 15% for a variable then just exclude the whole variable from the data frame and hence from the analysis, it is likely that the information it had still resides in the combination of the many other variables.

2) If you have less than 15% missing variables try to impute the values of the variable. In the extreme case you could apply multiple imputations and repeat the analysis on each imputation set and then pool the results. However for this exercise you can follow a simple strategy such as imputing the value with mean/median/mode. This would be a limitation but imputing missing values is not the focus of our exercise on causal inference.

```{r}

# => Inspect the proportion of missing values for each variable and follow the advice above.
print("===================================================")

#for every column (sapply) give True or False depending on whether it is na, then take the mean of that (treat True/False as 1/0)-> proportion of missing values for each variable.
missing.fraction <- sapply(rhc.small, function(column) mean(is.na(column)))
#print all variables that have missing values in descending order.
print("Variables that have missing values, sorted by descending proportion of missing values:")
print(sort(missing.fraction[missing.fraction > 0], decreasing = TRUE))

#take names of all the variables where the fraction is above the threshold (0.15) and drop those columns from the dataframe (keep all others). This got rid of sdld3p and urin1 which were actually the only two vars with missing values (see earlier printed list above), so the following imputation is redundant, nonetheless:
rhc.small <- rhc.small[ , !(names(rhc.small) %in% names(missing.fraction)[missing.fraction > 0.15])]
for (v in names(missing.fraction)[missing.fraction > 0 & missing.fraction <= 0.15]) {
  rhc.small[[v]][is.na(rhc.small[[v]])] <- median(rhc.small[[v]], na.rm = TRUE)
} #loops over variables that have a missing value proportion between 0 and 0.15 (excluding the former, including the latter), and impute the missing values for those variables as the median of the present values of that variable.
#We chose median as its generally more reliable and less skewed than average. And for data such as urin1 where the mode is 0, it offers a bit more information over just plugging the mode.


sprintf("There are %d missing values left in the rhc.small dataframe (should be 0)", sum(is.na(rhc.small)))
print("Sanity check that dimensions of rhc.small dataframe did not change:")
print(dim(rhc.small))

#Note: watch out because re-running this cell wont show the two dropped vars, since they no longer exist, do "run all" to see that
print("===================================================")
```
```{r}
# Check/see that all variables are numeric or int
str(rhc.small)
```
Now we are ready to start the analysis.

## Selecting variables to control for

According to the disjunctive cause criterion, we want to find variables (to control for) that are predictive of treatment and/or the outcome. In this assignment we will use the criterion to select all (pre-treatment) variables that are predictive of the outcome. This will include true confounders (that also affect the treatment, and risk factors that only affect the outcome). All our variables in rhc.small, except the outcome of course, are pre-treatment variables (none is measured AFTER the intervention) so they are all good candidates for the selection.

The truth is that the best way to select these variables is to use clinical knowledge. However, in this assignment we will do this in a data-driven and practical way. Specifically, we will operationalize "predictive of the outcome" as the following criterion: a variable is predictive of the outcome if its univariate association with the outcome is significant at the 0.1 level. Because our outcome is binary we can translate this criterion to: For every candidate variable in rhc.small fit a logistic regression model to predict outcome (outcome.died) using only the candidate variable as a predictor and retain the variable if the p-value of its association with the outcome is ≤ 0.1, and otherwise discard the variable. Note: Do not include the treatment in the selection. 

```{r}
print("===================================================")
# => Fit logistic regression models for each variable separately with the outcome and retain those with an association having a p-value of ≤ 0.1. Tip: you may want to automate the whole process!

#Why 0.1 and not 0.05? It's mainly because variable screening should have looser requirements than hypothesis testing because missing a possible confounder is much more harmfull (potentially increased bias significantly), compared to the additional cost of controlling for a variable that isn't actually a confounder.

#Make our candidates, but we first remove two variables the treatment and outcome, since checking if outcome affects outcome is pointless and we aren't experimenting on whether the treatment works, just looking for confounders, so ignoring those two:
candidates <- setdiff(names(rhc.small), c("treatment.swang", "outcome.died"))

p.values <- c()
#v is our placeholder, we iterate over each variable, plug it into the glm, the "family = binomial" flag sets it to logistic regression. summary(model)$coefficients gives the coefficient estimate, z-value, std error, and p-value. Index into it to extract p-value for variable coefficient.
for (v in candidates) {
  model <- glm(as.formula(paste("outcome.died ~", v)), family = binomial, data = rhc.small)
  p.values[v] <- summary(model)$coefficients[2, 4]
}
print(sort(p.values))


# => Put the names of the selected variables (without the treatment and outcome variables) in the variable called xvars by writing xvars <- c("ARF", ...) or do it automatically!
# => Print xvar

#filter out the variables with p-value <= 0.1 and put into xvars
xvars <- names(p.values)[p.values <= 0.1]

print("Variables with p-value <= 0.1:")
print(xvars)
sprintf("Number of fariables with p-value <= 0.1: %d", length(xvars))


# => Put these selected variables in a data.frame called rhc.selected that also includes the treatment and outcome variables to this data frame.
rhc.selected <- rhc.small[ , c(xvars, "treatment.swang", "outcome.died")] #Put everything back into the dataframe 
print("Dimensions of new dataframe with just selected variables:")
print(dim(rhc.selected))
# print("===================================================")
```


## Inspect balance before matching

## look at "table 1"
```{r}
print("===================================================")
#=> look at "table 1", complete the command by filling in the "..."
# CreateTableOne's strata argument tells it: "compute all these variable summaries (mean and stdev) separately for each level of this strata/grouping variable (treatment vs no treatment), then line them up side by side (plus the SMD between them). test=False ensures to ignore p-values since we want SMD and not significance tests that are sensitive to sample size, as discussed in the slides.
table1<- CreateTableOne(vars=xvars,strata="treatment.swang", data=rhc.selected, test=FALSE)
## include standardized mean difference (SMD)
print(table1, smd=TRUE)
## For which variables we do not have a good balance? Look at the slides for a guideline. 
# Expect to see several variables with high SMD score since the patients who got RHC and those who didn't were not randomized, so they likely differ on covariates like illnes, age, comborbidities etc. Here we quantify that: for each variable, is the distribution similar between the treated and not treated group? Then, after matching we can do this analysis again and see if it fixed some/all of the imbalances between groups, so that now the treatment does appear to be randomized across each covariate. 

# Following the slides:
smd_values = ExtractSmd(table1)
adequate_balance <- rownames(smd_values)[smd_values < 0.1]
not_too_alarming <- rownames(smd_values)[smd_values >= 0.1 & smd_values < 0.2]
imbalanced <- rownames(smd_values)[smd_values > 0.2]

sprintf("The following %d variables are adequately balanced between treatment and no treatment:", length(adequate_balance))
print(adequate_balance)

sprintf("The following %d variables are decently balanced between treatment and no treatment (on the border but not too alarming):", length(not_too_alarming))
print(not_too_alarming)

sprintf("The following %d variables are too imbalanced between treatment and no treatment:", length(imbalanced))
print(imbalanced)
print("===================================================")
```


## Greedy matching on Mahalanobis distance
Let us now do matching in the hope that the balance will get better so we can have a more valid inference.

```{r}
print("===================================================")

#Match() function has to randomly chose between equal best matches when replacement = False. Thus, set seed for reproducibility of results.
set.seed(42)


# We want to match the treated with the untreated subjects. Find 1 on 1 match.
# # => Complete the command
#greedymatch <- Match(Tr=,M=, X=rhc.selected[xvars], replace=FALSE) 
# read the documentation for Match() by typing ?Match.
# => What does replace=FALSE mean?
# => What is inside X in the command above?

# Tr is a vector indicating for each observation whether it belongs to the treatment or not treated group, M determines how many control instances we match to each treatment instance (here 1:1 matching) and X is a matrix (therefore added as.matrix, per the documentation) with all the variables (covariates) to match on (calculate the Magalanobis distance with). replace = False means matching without replacement, i.e. once an untreated subject has been used (matched), it can't be matched again against another treated subject. This forces each matched control to be unique, so the order of matching matters, and thus this typically increases bias relative to matching with replacement. If we match with replacement, the same "good" control can be matched against multiple treated subjects, which gives closer matches on average (lower bias), but increases variances, highlighting the inherent bias-variance tradeoff. Lastly, Weight = 2 specifies to use the true Mahalanobis distance as distance metric for matching, including covariance correction (to ensure to not double penalize difference along two correlated dimensions as if they were independent), rather than only standardizing by variance which is what the default inverse-variance-weighting does.
greedymatch <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = as.matrix(rhc.selected[xvars]), 
                      replace = FALSE, Weight = 2)

print("===================================================")
```

```{r}
print("===================================================")
greedymatch$index.treated # => What do you think this is?
greedymatch$index.control # => What do you think this is?
# greedymatch$index.treated is a vector containing for all treated observation in the matched dataset, theit observation (index/row) numbers in the original dataset (rhc.selected). greedymatch$index.control contains the same indices, but then for the control subjects in the matched dataset (those selected during matching). Notably, index.treated[i] and index.control[i] together identify the i-th matched pair: row index.treated[i] is a treated subject, and row index.control[i] is the untreated subject matched to them. So this means that these two vectors let you reconstruct exactly which original rows ended up in the matched dataset, and how they're paired together, e.g., rhc.selected[greedymatch$index.treated[1],] and rhc.selected[greedymatch$index.control[1], ] give you the first matched pair. Lastly, note that since M=1 and replace=FALSE, index.control should contain just as many values as index.treated and no repeated values (each control used at most once), while each treated subject is matched since no caliper was specified.
print("===================================================")
```


=> Are all treated subjects matched to a control one?

```{r}
matched <- rhc.selected[unlist(greedymatch[c("index.treated","index.control")]), ]
matched[1,]  # This is the first treated subject
print("===================================================")
matched[length(greedymatch$index.treated)+1,] # => What do you think this subject is?
# Because unlist(greedymatch[c("index.treated","index.control")]) concatenates all of index.treated first, then all of index.control, the resulting matched dataframe contains in rows 1 to length(index.treated) → all the treated subjects in matching order, and in rows length(index.treated)+1 to 2×length(index.treated) (since 1:1 matching) all the matched controls, in the same order as their treated counterparts. Therefore, matched[length(greedymatch$index.treated)+1,] is the control subject matched to the first treated subject (the one printed just above here).
print("===================================================")
```

=> Added analysis about number of matched patients.

```{r}
num_matched_treated = length(greedymatch$index.treated)
num_matched_control = length(greedymatch$index.control)
sprintf("Number of matched treated patients is %d, and the number of matched control patients is %d", num_matched_treated, num_matched_control)
print("Number of matched treated and control patients should be equal since 1:1 matching.")

num_not_matched_treated = num_treated - num_matched_treated
fraction_treated_matched = num_matched_treated/num_treated*100
num_not_matched_control = num_not_treated - num_matched_control
fraction_control_matched = num_matched_control/num_not_treated*100


sprintf("There are %d treated patients not matched, meaning that %.2f percent of treated patients got matched",num_not_matched_treated, fraction_treated_matched)
sprintf("There are %d control patients not matched, meaning that %.2f percent of control patients got matched",num_not_matched_control, fraction_control_matched)
print("===================================================")
```


Get table 1 for matched data with standardized differences
```{r}
print("===================================================")
# CreateTableOne's strata argument tells it: "compute all these variable summaries (mean and stdev) separately for each level of this strata/grouping variable (treatment vs no treatment), then line them up side by side (plus the SMD between them). test=False ensures to ignore p-values since we want SMD and not significance tests that are sensitive to sample size, as discussed in the slides.
matchedtab1 <- CreateTableOne(vars=xvars,strata="treatment.swang", data=matched, test=FALSE) # => Complete the command to create table 1 after matching
# => Print the balance with smd. 
print(matchedtab1, smd=TRUE)

  
# Expect to see less variables with high SMD scores after matching. 

# Following the slides:
smd_values_matched = ExtractSmd(matchedtab1)
adequate_balance_matched <- rownames(smd_values_matched)[smd_values_matched < 0.1]
not_too_alarming_matched <- rownames(smd_values_matched)[smd_values_matched >= 0.1 & smd_values_matched < 0.2]
imbalanced_matched <- rownames(smd_values_matched)[smd_values_matched > 0.2]

sprintf("The following %d variables are adequately balanced between treatment and no treatment:", length(adequate_balance_matched))
print(adequate_balance_matched)

sprintf("The following %d variables are decently balanced between treatment and no treatment (on the border but not too alarming):", length(not_too_alarming_matched))
print(not_too_alarming_matched)

sprintf("The following %d variables are too imbalanced between treatment and no treatment:", length(imbalanced_matched))
print(imbalanced_matched)

# => After matching how many "bad" variables with SMD > 0.2 are there?
print("After matching there are 7 bad variables with SMD > 0.2, compared to 12 before matching.")


# print("===================================================")
```

## Outcome analysis
```{r}
print("===================================================")
#n <- length(greedymatch$index.treated) #n = number of treated and control subjects in matched
#y_trt <- matched$outcome.died[1:n]# => Put here the vector of outcomes y (from matched$outcome.died) of those in the treatment group
#y_con <- matched$outcome.died[(n+1):(2*n)]# Put here the vector of outcomes y (from matched$outcome.died) of those in the control group

y_trt <-matched$outcome.died[matched$treatment.swang == 1]# => Put here the vector of outcomes y (from matched$outcome.died) of those in the treatment group
y_con <- matched$outcome.died[matched$treatment.swang == 0]# Put here the vector of outcomes y (from matched$outcome.died) of those in the control group
print("===================================================")
```

## Let's do a paired t-test to see if there are differences in death between the matched groups. This is not the best test because the data is binary but we want to just get a feeling of the differences.

```{r}
diffy <- y_trt-y_con # pairwise difference
t.test(diffy)
```

=> Look at the p-value. Is it below 0.05?
Yes, the p-value = 1.453e-11, so far below 0.05. 
=> What does this mean?
This means the mean difference in outcome (death) between treated and not treated groups (0.0911) is significantly different from 0. In other words, RHC patients-treated patients had significantly higher death rates than their matched controls. Note however that a t-test isn't ideal here since the binary data violates the normality assumption of a t-test.

## We want to use the McNemar test, which is the best choice for paired (matched) binary data
```{r}
table(y_trt,y_con)
# In this table you will see the 2x2 table:  
#         y_con
# ytreat     0     1
#       0    a     b
#       1    c     d

# Enter these a, b, c, and d numbers in the following McNemar test
# print("===================================================")
#mcnemar.test(matrix(c(),2,2)) # => Complete the command by filling the a, b, c, and d numbers in the correct order.
mcnemar.test(table(y_trt,y_con))
# => Is the p-value less than 0.05? What does this result mean?
#Yes, the p-value is 2.294e-11, much less than 0.05. This means the discordant pairs (b and c, where the outcome of treated and control disagreed) are imbalance, i.e., treatment status is associated with chance of death within matched pairs (evidence of treatment effect). Here, since c=538 is statistically significantly greater than b=339, there seems to be evidence that there is a higher chance of death among the treated patients within matched pairs, which is consistent with the t-test above.
# print("===================================================")
```

## The last way we want to analyse the outcome is with logistic regression
```{r}
gmodel1 <- glm(outcome.died ~ treatment.swang, family=binomial, data=matched)
print("===================================================")
# => Is the coefficient of treatment significant? you can use summary(gmodel1) or confint(gmodel1)
summary(gmodel1)
# In summary(gmodel1) we can see that the p-value for treatment is 4.34e-10, indicating that the coefficient of treatment is significant. 

# =>  What is the odds ratio that is associated with treatment? You can easily derive it from the coefficient (Google otherwise how to get the odds ratio from the coefficient)
sprintf("Coefficient of treatment is: %.5f.", coef(gmodel1)["treatment.swang"])
# The estimated treatment coefficient is 0.39461, thus the odds ratio associated with treatment is exp(0.39461) = 1.4838. This means that the odds of death increase by a factor of 1.4838 in the treatment group compared to the no treatment group.

# =>  If the coefficient is positive, what does that mean in terms of odds ratio?
# As explained above, if the coefficient is positive the odds ratio (exp(coefficient)) will be > 1, meaning that the treatment group has a higher chance of death compared to the not treated group.

# => If the coefficient is negative, what does that mean in terms of odds ratio?
# Meanwhile, if the coefficient is negative the odds ratio (exp(coefficient)) will be < 1, meaning that the treatment group has a lower chance of death compared to the not treated group.

# The great thing about modelling is that we can now again adjust for the variables with the worst SMD. This is called doubly-robust estimation. Adjust for the variable with the worst SMD that you encountered before. To adjust you just add the name of the variable in the logistic regression formula. => Complete the command
# aps1 is the variable with the worst SMD (0.501) after matching, thus we adjust for it in logistic regression:
gmodel2 <- glm(outcome.died ~ treatment.swang + aps1, family=binomial, data=matched)
summary(gmodel2)
# => What is the odds ratio for treatment now?
# In summary(gmodel2) we can see that the p-value for treatment is 2.57e-2, indicating that the coefficient of treatment is still significant (< 0.05). 
sprintf("Coefficient of treatment is: %.5f.", coef(gmodel2)["treatment.swang"])
# The estimated treatment coefficient is 0.19891, thus the odds ratio associated with treatment is exp(0.19891) = 1.2201. This means that the odds of death increase by a factor of 1.2201 in the treatment group compared to the no treatment group. Notably, this is a considerably smaller odds ratio than before adjusting for aps1 (note that the p-value is also a lot larger now (2.57e-2 vs 4.34e-10)).
# print("===================================================")
```


## Propensity score matching

Fit a propensity score model to predict treatment (not the outcome!). Use logistic regression with glm
```{r}
print("===================================================")
# now the outcome variable is treatment (treatment.swang), not outcome.died since we're modeling "who was likely to receive RHC, given their covariates". "~ ." includes all other variables as predictors (besides what we're modeling).
psmodel <- glm(treatment.swang ~ ., family = binomial, data = rhc.selected[, c(xvars, "treatment.swang")])# => Use the variables we found before in xvars in the first assignment to predict treatment

summary(psmodel)   # show coefficients etc
pscore<-psmodel$fitted.values # create propensity score

print("===================================================")
```

# Now it is time to check the positivity assumption
# let's check the overlap between the propensity scores for the two groups. We do that before the matching.
```{r}
print("===================================================")
#  => plot the density of the propensity scores of the treated and untreated. Tip: you might want to use par(mfrow=c(2,1)) to create 2 panels (2 rows, 1 column) and then plotting the first hist (histogram), with say, ylim=c(0,650) and col='darkblue', and the second histogram with ylim=c(650, 0) (note that we flipped the order) and col='darkblue'. This will plot the blue histogram as a normal histogram and the red one will be plotted upside down underneath the blue one so you can easily see the overlap.


brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "Propensity score distribution in treated vs untreated groups.", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap them nicely
hist(pscore[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlapping histogram

# => What do you think about the overlap? 
# Across most of the range of propensity scores there is substantial overlap (between 0.1 and 0.8), there exist comparable treated and untreated subjects to match at most points along the score. Only below 0.1 and between 0.8-0.9, there is little overlap, indicating a possible positivity violation in those small regions, which risks no valid comparison and thus few matches in those regions (since we will be using a caliper to ensure we avoid poor, distant matches). However, since this is real RHC data, we expected this decent but imperfect overlap in the tails, since, as we observed before, there simply are significantly more deaths (and thus higher propensity scores) in the treated group. In conclusion, the positivity assumption is okay here, but we note a possible limitation in the tails.


print("===================================================")
```

We want to do greedy matching on PS using Match with a caliper. Note that instead of the probability PS itself, we usually use the log odds (logit) of PS which is log(PS/(1-PS)) which is also equal to log(PS)-log(1-PS) . So let us first write a function (of one line) to calculate the logit.

```{r}
print("===================================================")
logit <- function(p) {log(p/(1-p))} # => Complete this function to calculate the log odds
print("===================================================")
```


Now we want to use Match() with rhc.selected$treatment.swang as treatement, with 1 matched control per treated subject, with X as the logit(pscore), with no replacement and with a caliper of 0.2

```{r}
print("===================================================")
#Match() function has to randomly chose between equal best matches when replacement = False. Note that this is much more likely here than before since now we are only matching on 1 variable (propensity) rather than all the variables in xvars. Thus, set seed for reproducibility of results.
set.seed(42)

# Compute all logits
logits <- sapply(pscore, logit)

#psmatch <- Match(Tr=mydata$treatment,...) # => complete the command

psmatch <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logits, 
                      replace = FALSE, Weight = 2, caliper = 0.2)
print("===================================================")
matched_ps <- rhc.selected[unlist(psmatch[c("index.treated","index.control")]), ]
```


## Get standardized differences. Note: xvars is the same as in the first assignment.

```{r}
print("===================================================")
matchedtab1_ps <- CreateTableOne(vars=xvars, strata ="treatment.swang", 
                            data=matched_ps, test = FALSE) # => Complete the command
# => print the matching table showing the SMD
print(matchedtab1_ps, smd=TRUE)

# Expect to see less variables with high SMD scores after matching. 

# Following the slides:
smd_values_matched_ps = ExtractSmd(matchedtab1_ps)
adequate_balance_matched_ps <- rownames(smd_values_matched_ps)[smd_values_matched_ps < 0.1]
not_too_alarming_matched_ps <- rownames(smd_values_matched_ps)[smd_values_matched_ps >= 0.1 & smd_values_matched_ps < 0.2]
imbalanced_matched_ps <- rownames(smd_values_matched_ps)[smd_values_matched_ps > 0.2]


sprintf("Max SMD: %.3f", max(smd_values_matched_ps))

sprintf("The following %d variables are adequately balanced between treatment and no treatment:", length(adequate_balance_matched_ps))
print(adequate_balance_matched_ps)

sprintf("The following %d variables are decently balanced between treatment and no treatment (on the border but not too alarming):", length(not_too_alarming_matched_ps))
print(not_too_alarming_matched_ps)

sprintf("The following %d variables are too imbalanced between treatment and no treatment:", length(imbalanced_matched_ps))
print(imbalanced_matched_ps)

# => Are there any variables with bad balance now?
print("After matching there are 0 bad variables with SMD > 0.2 (even 0 with SMD > 0.1), compared to 7 after covariate matching and 12 before matching.")
print("===================================================")
```

## Outcome analysis after propensity score matching
```{r}
print("===================================================")
y_trt.ps <- matched_ps$outcome.died[matched_ps$treatment.swang == 1] # => Put here the vector of outcomes y (from matched$outcome.died) of those in the treatment group
y_con.ps <- matched_ps$outcome.died[matched_ps$treatment.swang == 0]# Put here the vector of outcomes y (from matched$outcome.died) of those in the control group
print("===================================================")
```


Perform a t-test
```{r}
# Perform paired t-test. Just to get a feeling of the result
diffy.ps <- y_trt.ps - y_con.ps # pairwise difference
t.test(diffy.ps)
print("===================================================")
# => Is the test significant?
#Yes, the p-value = 9.767e-06, so well below 0.05. 
# This means the mean difference in outcome (death) between treated and not treated groups (0.07151096) is significantly different from 0. In other words, RHC patients-treated patients had significantly higher death rates than their matched controls. Note however that a t-test isn't ideal here since the binary data violates the normality assumption of a t-test.
print("===================================================")
```


Use the McNemar test as before. Enter the a, b, c, and d numbers you get from the 2x2 table in the McNemar test
```{r}
print("===================================================")

table(y_trt.ps,y_con.ps)

#mcnemar.test(matrix(c(...),2,2)) # => Complete the command
# => Is the p-value less than 0.05? What does this result mean?
# => Compare the difference in the means before and after matching.


# In this table you will see the 2x2 table:  
#         y_con.p
# ytreat.p     0     1
#       0    a     b
#       1    c     d

mcnemar.test(table(y_trt.ps,y_con.ps))
# => Is the p-value less than 0.05? What does this result mean?
#Yes, the p-value is 1.208e-05, a lot less than 0.05. This means the discordant pairs (b and c, where the outcome of treated and control disagreed) are imbalance, i.e., treatment status is associated with chance of death within matched pairs (evidence of treatment effect). Here, since c=457 is statistically significantly greater than b=333, there seems to be evidence that there is a higher chance of death among the treated patients within matched pairs, which is consistent with the t-test above.
# => Compare the difference in the means before and after matching.
# Difference in means y_treat - y_con after matching on propensity is 0.07151096  (see t-test right above).
# Difference in means y_treat - y_con after matching on covariates is 0.09111722  (see first t-test).
# Difference in means y_treat - y_con before matching is 0.6804 - 0.6297 (from original descriptive data analysis at the top) = 0.0507. 
# Since propensity matching achieved better covariate balance (lower SMDs), it more successfully removed confounding on the measured covariates, so it's the more trustworthy estimate of the two. That the raw difference in means estimate happens to resemble the propensity-matched estimate is likely coincidental sample-specific behavior, and not proof that confounding was absent. The SMD tables tell us that propensity matching demonstrably balanced the covariates better, which is why its estimate is more trustworthy. The Mahalanobis number being even slightly higher is more likely an artifact of incomplete balancing (and/or a different matched subsample) than a genuine midpoint in some meaningful confounding-removal trajectory. Anyways, the difference in means after balancing the matches well in the propensity-based approach, indicate that after adjusting for those possible confounders, there is still a difference in means between the treated and not treated groups, which is in line with the significance found in the t-test, McNemar test and logistic regression below.

print("===================================================")
```

Use logistic regression
```{r}
gmodel3 <- glm(outcome.died ~ treatment.swang, family=binomial, data=matched_ps)
print("===================================================")
# => What is the odds ratio of treatment?
summary(gmodel3)
# In summary(gmodel3) we can see that the p-value for treatment is 0.000189, indicating that the coefficient of treatment is significant. 
sprintf("Coefficient of treatment is: %.5f.", coef(gmodel3)["treatment.swang"])
# The estimated treatment coefficient is 0.26655, thus the odds ratio associated with treatment is exp(0.26655) = 1.3055. This means that the odds of death increase by a factor of 1.3055 in the treatment group compared to the no treatment group.


# => Now correct also for the least well balanced variable
# scoma1 is the variable with the worst SMD (0.037) after matching, thus we adjust for it in logistic regression:
gmodel4 <- glm(outcome.died ~ treatment.swang + scoma1, family=binomial, data=matched_ps)
# => What is the odds ratio of treatment now?
summary(gmodel4)
# In summary(gmodel4) we can see that the p-value for treatment is 0.000163, indicating that the coefficient of treatment is still significant (<0.05). 
sprintf("Coefficient of treatment is: %.5f.", coef(gmodel4)["treatment.swang"])
# The estimated treatment coefficient is 0.27114, thus the odds ratio associated with treatment is exp(0.27114) = 1.3115 This means that the odds of death increase by a factor of 1.3115 in the treatment group compared to the no treatment group. Notably, this is a slightly larger odds ratio than before adjusting for scoma1, highlighting the fact that even the variable with the highest SMD value was still not significantly confounding the results of gmodel3.

print("===================================================")
```


# Open part (30% of the grade)
Instead of the glm to calculate the propensity score use any machine learning technique you like such as a regression tree (using rpart), or random forests (randomForest) or neural networks (neuralnet) or another formalism in R to create the propensity score model and to predict the probabilities on such a model. Note that you might need to tweak the model parameters in order to improve it. Compare your results to those you obtained by matching (via the Mahalanobis distance and the propensity score with logistic regression). Which approach do you prefer, and why?

```{r}
#Installing and loading required packages. Can comment installation lines out after ran once.
install.packages("randomForest")
install.packages("caret")
install.packages("data.table") # reinstall into the user library: the system copy was built for an older R and fails to load (undefined symbol SETLENGTH), which breaks library(caret)
install.packages("pROC")
install.packages("CBPS")
install.packages("ggplot2")
install.packages("nnet")
library(CBPS)
library(randomForest)
library(pROC)
library(caret)
library(ggplot2)
library(nnet)
```


So far we have compared the Mahalanobis distance and propensity score, with the latter coming in as the strong favourite due to its more comprehensive nature - in the end encoding within it more information than the euclidian distance matching could. However we still have the case of machine learning models. As a group we have decided to try multiple approaches and compare our findings and results between them, looking at different performance metrics and characteristics. 


We decided to compare a random forest (RF), a neural network, and covariate balanced propensity score (CBPS) against the standard glm because each represents a genuinely different strategy for estimating the propensity score, and together they probe different weaknesses of the baseline logistic model. The glm assumes a strictly additive relationship, that is linear in the log odds, between covariates and treatment — a strong, potentially wrong assumption we take simply based on faith. Random forest relaxes this assumption by learning through hierarchical thresholds rather than smooth linear terms, closer to how clinical decisions are actually made (a lab value crossing a cutoff, not a proportional risk score), and is able to capture nonlinearities and interactions that the standard glm isn't able to express. On the otherhand, the neural network gives a different kind of flexibility again; it is a smooth, continuous function approximator that can, in principle, capture arbitrary nonlinear relationships (including interactions) without the discrete splits a tree is restricted to. However, as a result, it is far less interpretable and more sensitive to tuning and preprocessing. CBPS (Imai & Ratkovic, 2014) is different from the other two: rather than adding flexibility to the functional form, it keeps the same linear logistic form as the glm but changes directly what the model is optimized for (adds constraints to the loss). This explicitly targets covariate balance in its fitting procedure rather than optimizing predictive likelihood alone. Testing all four side by side lets us separate distinct questions that are usually mixed up: does a more flexible/powerful functional form produce a better propensity score (RF, NN vs. glm), if so, does that produce better balanced covariate groups, and does optimizing directly for the right objective matter more than model flexibility at all (CBPS vs. glm)?

As a first optimization step and comparison between models, we will be finetuning some of the hyperparameters of the RF and NN models to optimize for covariate balance. Very important here is to emphasize the difference between propensity score predictive power and ability to create well-balanced covariate groups. The actual downstream of the propensity scores is to use them to create well balanced groups, not to perfectly classify each patient as treated/not treated. These two goals are related but not the same, and can actively conflict (e.g. a model that predicts treatment too well tends to push scores toward 0 and 1 (near-certain predictions), which destroys overlap/positivity). Maybe a model that is (slightly) overfitted balances better because it moves the propensity scores away from 0/1 and more towards the middle values which we results in more overlap and better positivity. Therefore, we optimize hyperparameters based on how well they balance the covariates by for each combination of hyperparameters look at the max SMD score, and the number of variables that have below 0.1, between 0.1-0.2 and over 0.2 SMD scores respectively.Note that for the baseline GLM and CBPS models there is no hyperparameter to do, since they essentially have fixed capacity once the formula/functional form is fixed. We will, however, investigate the addition of interaction terms in those two models. 

First up is Random Forest. It was primarily chosen due to its simplicity; it is often the classic first choice when ML is in the conversation, it may not have stellar results compared to other methods but it is steady, reliable and relatively interpretable. An interesting concept that came to mind (only a theory at this point, yet to be proven or disproven later in this chapter) is that of continuous vs discrete values. You see our previous two metrics, particularly PS is a continuous value, which works perfectly fine, but in the medical field our decisions are rarely motivated by proportions, but rather by thresholds. A patient is not "62% hypertensive", but rather a certain blood pressure threshold triggers treatment. Clinical reasoning can be observed as a sequence of cutoff points, which coincidentally is the precise mechanism a decision tree, and by extension - Random Forest, operates with.
Whether this structural resemblance translates into better estimates, is what we set out to test in this next section.

```{r}
rf.data <- rhc.selected[, c(xvars, "treatment.swang")]
rf.data$treatment.swang <- as.factor(rf.data$treatment.swang)
set.seed(42) #standard seed initialization for reproducibility

```

With the base data set up, its time for our hyperparameters and model itself

```{r}
rf.model <- randomForest(treatment.swang ~ ., data = rf.data, ntree = 500, nodesize = 25)
print(rf.model)
plot(rf.model)

```
We started with the default of 500 trees in the RF. Adding more trees never hurts the result, but adds compute time, and also by plotting rf.model we can see that the probabilities even out and are stable well before the 500 mark, suggesting that upping the number is not necessary. As such, from here on we only use 300 trees.
The two main diasl we have are node size and the number of variables randomly considered as candidates at each split. Node size is the minimum number of patients allowed in a leaf. Its default is 1, meaning a tree keeps splitting until every leaf is pure, so in the limit each leaf is essentially one patient and the tree has memorized the training data. Our starting idea was that this would be bad for a propensity score: a leaf that has memorized a single patient can only ever say "treated" or "not treated", so we expected the scores to pile up at 0 and 1, which would give infinite logits and destroy the overlap we need for matching. Forcing larger leaves seemed like the natural fix, since a leaf of 25 patients has to base its vote on a mix of treated and untreated cases rather than on a single one, and we expected that to produce smoother scores and better overlap. We therefore chose node size = 25 as a starting point, large enough to stop memorization but small enough that the trees can still tell patients apart. Secondly, the default for the number of variables randomly considered as candidates at each split is sqrt(number of predictors), so ~ sqrt(40) which is roughly 6 here. A lower number of possible variables increases randomness and thus decorrelation between trees, which makes them less prone to overfitting but individually have weaker predictive power. A larger number of possible variables make the trees more similar to each other and more powerful, but also more prone to overfitting. To test this reasoning and find out which values are optimal for balancing, we test it. 


For each of several node size values and number of randomly selected candidate variables we fit a forest, take its propensity scores, and run exactly the same matching and balance check we used for the glm propensity score, so that node size is judged by the thing we actually care about: overlap in the histograms and balance after matching.


```{r}
nodesizes <- c(1, 5, 10, 25, 50, 100)
mtrys <- c(2, 3, 4, 6, 10, 15)  # default for classification is floor(sqrt(n_vars) ~= 6 here, so spread around it but not useful to go lower than 3 since than you get almost entirely random splits in the trees.

ps.by.combo <- list()
nodesize.comparison <- data.frame()

for (ns in nodesizes) {
  for (mt.val in mtrys) {
    set.seed(42) #Makes the forest's randomness (e.g. bootstrap sampling and feature selection at splits) reproducible
    m <- randomForest(treatment.swang ~ ., data = rf.data, ntree = 300, nodesize = ns, mtry = mt.val) #Fits forest for current nodesize/mtry combo
    ps <- predict(m, type = "prob")[, "1"] #Propensity score = fraction of trees voting "treated"
    
    key <- paste0("ns", ns, "_mtry", mt.val) #Unique key per combination, e.g. "ns10_mtry6"
    ps.by.combo[[key]] <- ps #Store propensity scores keyed by combo

    # Scores of exactly 0 or 1 have an infinite logit, so we push them up to 0.005 or down to 0.995 to not get an error
    ps.c <- pmin(pmax(ps, 0.005), 0.995)
    set.seed(42)  #TODO: is this needed again?
    # 1:1 greedy matching with Mahalanobis distance on the logit transformed propensity score predicted by the RF, with no replacement and a caliper of      0.2 as before so that we can accurately compare the propensity score calculation method (controlling for other hyperparameters). 
    mt <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logit(ps.c),
                replace = FALSE, Weight = 2, caliper = 0.2)
    md <- rhc.selected[unlist(mt[c("index.treated", "index.control")]), ] #Extract matched data as before
    smd <- ExtractSmd(CreateTableOne(vars = xvars, strata = "treatment.swang", data = md, test = FALSE)) #Calculate SMD scores to check balancing

    # Comparison table; add 1 row of info per hyperparam combo
    nodesize.comparison <- rbind(nodesize.comparison, data.frame(
      nodesize       = ns,
      mtry           = mt.val,
      oob.error      = round(tail(m$err.rate[, "OOB"], 1), 3), #Out-of-bound error rate, i.e. held-out validation error for each full RF of 500 trees (similary to classical cross validation). Tells us how accurately predictive model is, but note that we don't want model to predict propensity scores perfectly as 0/1 since then there is bad overlap/positivity in scores. 
      extreme.scores = sum(ps < 0.01 | ps > 0.99), # patients with almost certain predictions, don't want this high since leads to unbalanced dataset (bad for positivity)
      matched.pairs  = length(mt$index.treated), # how many subjects succesfully found a caliper satisfying match (fewer = worse overlap/positivity)
      max.smd        = round(max(smd), 3), #maximum found smd (most unbalanced variable)
      n.smd.under.0.1 = sum(smd < 0.1), #number of variables well-balanced
      n.smd.0.1.0.2  = sum(smd >= 0.1 & smd <= 0.2), #number of variables in danger zone
      n.smd.over.0.2 = sum(smd > 0.2))) #number of variables not well-balanced
  }
}
# sorted so the best-balancing combos appear first
print("Best balanced variables by max SMD value:")
print(nodesize.comparison[order(-nodesize.comparison$max.smd), ]) 
print("Best balanced variables by number of variables well-balanced:")
print(nodesize.comparison[order(-nodesize.comparison$n.smd.under.0.1), ])

# heatmap of the balance metric to show balance results: number of variables well balanced (SMD < 0.1) and max SMD value.
ggplot(nodesize.comparison, aes(x = factor(nodesize), y = factor(mtry), fill = max.smd)) +
  geom_tile() +
  geom_text(aes(label = max.smd), size = 3) +
  scale_fill_gradient(low = "darkgreen", high = "darkred") +
  labs(x = "nodesize", y = "mtry", fill = "max SMD",
       title = "RF balance (max SMD) across nodesize x mtry grid") +
  theme_minimal()


ggplot(nodesize.comparison, aes(x = factor(nodesize), y = factor(mtry), fill = n.smd.under.0.1)) +
  geom_tile() +
  geom_text(aes(label = n.smd.under.0.1), size = 3) +
  scale_fill_gradient(low = "darkred", high = "darkgreen") +
  labs(x = "nodesize", y = "mtry", fill = "# SMD < 0.1",
       title = "RF balance (variables with SMD < 0.1) across nodesize x mtry grid") +
  theme_minimal()
```

These heatmaps tell a different story from the one we expected. For the node size, small values (1, 5 and 10) look best: one hump per group, both spread across roughly 0 to 0.9, and wide overlap in the middle. Larger values make things worse, and at 50 and 100 the untreated patients collapse into a spike near 0. The reason is that each tree votes for the majority class of its leaf, and with leaves of 100 patients almost every leaf is majority-untreated (62% of all patients in the entire dataset are), so the majority of trees vote "not treated" for most patients. Our worry about the forests remembering patients applied to a single tree, but the propensity score is the average vote of 500 trees grown on different bootstrap samples and different random subsets of features, and that averaging combined with the randomness already smooths the scores sufficiently. Looking at the table - nodesize 10 and 25 give the best balance (2 variables with SMD above 0.1, max SMD 0.113 and 0.115), while 50 and 100 produce many more near-certain scores (45 and 128) and worse balance (4 and 8 variables above 0.1). Nodesize 1 is slightly worse than 10 as well, so the very smallest leaves add a little noise. Between 10 and 25, nodesize 10 has fewer extreme scores (4 versus 20) and keeps more matched pairs (1672 versus 1656), so that is what we use for the final model; our starting guess of 25 was close, but not the best. Note already that even the best forest leaves two variables above 0.1, where the glm propensity score left none, a point we shall return to. Overall the choice between node count is narrower than the graphs tell at first. #TODO update paragraph. select best hyperparams that score best on those metrics but also still have high number of matched pairs since that indicates overlap/positivity (and then also show this is the case in an overlap graph)



Next up is the neural network. As discussed before, random forests reason through discrete, threshold-based splits, while a a neural network offers the opposite kind of flexibility: a smooth, continuous function of the covariates, fit through gradient-based optimization rather than recursive partitioning. Since this let's us capture nonlinear relationships and interactions between covariates without being restricted to axis-aligned splits, we expect this to be able to optimize any goal relatively well. However, since this is a very powerful model, we need to ensure careful preprocessing (standardizing  numerically encoded inputs) and use regularization to avoid overfitting on a such a tiny dataset. As with random forest, we tune the most important hyperparameters: capacity (hidden layer size) and regularization strength (weight decay). However, again, following our own critique of predictive-accuracy-driven tuning, we select them based on the balance they produce after matching, not on how well they predict treatment assignment.




```{r}
# Preprocessing  (RF didn't need this)
nn.data <- rhc.selected[, c(xvars, "treatment.swang")]
nn.data$treatment.swang <- factor(nn.data$treatment.swang, labels = c("no", "yes"))  # nnet-friendly labels

# model.matrix handles factor -> one-hot expansion automatically and -1 drops intercept column
X.raw <- model.matrix(~ . - 1, data = nn.data[, xvars])

# Standardize (mean 0, sd 1): essential for NN training stability, unlike RF/glm
preproc <- preProcess(X.raw, method = c("center", "scale"))
X.scaled <- predict(preproc, X.raw)

nn.df <- data.frame(X.scaled, treatment.swang = nn.data$treatment.swang)

grid <- expand.grid(size  = c(2, 4, 6, 8, 12),            # number of hidden units
                     decay = c(0, 0.001, 0.01, 0.1, 0.5))  # L2 regularization strength

ps.by.combo.nn <- list()
nn.comparison <- data.frame()

for (i in 1:nrow(grid)) { #loop over grid of hyperparams
  sz <- grid$size[i]; dc <- grid$decay[i]
  set.seed(42) #seed for reproducibility of random intial weights etc.
  # Fit an NN for a max of 300 epochs on this combination of hidden size / decay values. trace = FALSE removes printing at each epoch. linout = FALSE makes it output probability (through sigmoid) rather than raw logit.
  m <- nnet(treatment.swang ~ ., data = nn.df, size = sz, decay = dc,
            trace = FALSE, maxit = 300, linout = FALSE) 
  ps <- predict(m, newdata = nn.df, type = "raw")[, 1] #propensity scores for each patient. Note that we just make it predict on full training data and we don't care so much about overfitting, since, as explained before, we just watch propensity scores that balance well, we don't care about if they over or underfit the propensity prediction a bit in the end. 

  key <- paste0("size", sz, "_decay", dc)
  ps.by.combo.nn[[key]] <- ps

  # Scores of exactly 0 or 1 have an infinite logit, so we clip them
  ps.c <- pmin(pmax(ps, 0.005), 0.995)
  set.seed(42)
  #Identical matching and SMD calculation steps as before
  mt <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logit(ps.c),
              replace = FALSE, Weight = 2, caliper = 0.2)
  md <- rhc.selected[unlist(mt[c("index.treated", "index.control")]), ]
  smd <- ExtractSmd(CreateTableOne(vars = xvars, strata = "treatment.swang", data = md, test = FALSE))

  # record hyperparam combination with same metrics as before
  nn.comparison <- rbind(nn.comparison, data.frame(
    size            = sz,
    decay           = dc,
    extreme.scores  = sum(ps < 0.01 | ps > 0.99),
    matched.pairs   = length(mt$index.treated),
    max.smd         = round(max(smd), 3),
    n.smd.under.0.1 = sum(smd < 0.1), 
    n.smd.0.1.0.2  = sum(smd >= 0.1 & smd <= 0.2), 
    n.smd.over.0.2 = sum(smd > 0.2)))
}

# 
print("Best balanced combos by max SMD value:")
print(nn.comparison[order(nn.comparison$max.smd), ])   #ascending: lowest (best) max SMD first

print("Best balanced combos by number of variables well-balanced:")
print(nn.comparison[order(-nn.comparison$n.smd.under.0.1), ])  #descending: most (best) well-balanced vars first

# Again, heatmaps showing the balancing of different hyperparam combinations.
ggplot(nn.comparison, aes(x = factor(size), y = factor(decay), fill = max.smd)) +
  geom_tile() +
  geom_text(aes(label = max.smd), size = 3) +
  scale_fill_gradient(low = "darkgreen", high = "darkred") +
  labs(x = "size", y = "decay", fill = "max SMD",
       title = "NN balance (max SMD) across size x decay grid") +
  theme_minimal()

ggplot(nn.comparison, aes(x = factor(size), y = factor(decay), fill = n.smd.under.0.1)) +
  geom_tile() +
  geom_text(aes(label = n.smd.under.0.1), size = 3) +
  scale_fill_gradient(low = "darkred", high = "darkgreen") +
  labs(x = "size", y = "decay", fill = "# SMD < 0.1",
       title = "NN balance (variables with SMD < 0.1) across size x decay grid") +
  theme_minimal()
```
#TODO update paragraph. select best hyperparams that score best on those metrics but also still have high number of matched pairs since that indicates overlap/positivity (and then also show this is the case in an overlap graph)



Lastly, we get to the CBPS model. As explained before, the actual goal is to create well-balanced treated vs control groups. The regular glm does not directly optimize for this since it simply fits the model to maximize the likelihood of correctly predicting treatment. Balance is never part of the fitting objective and is only measured post-hoc by checking SMD. CBPS (Imai & Ratkovic, 2014) fix this by changing directly what the model is optimizing.

Instead of fitting the propensity model purely by maximum likelihood, the propensity score can be estimated in a way that maximizes a combination of covariate balance as well as the prediction accuracy. CBPS uses the same parametric logistic form (β·X) as ordinary logistic regression, but chooses β by an augmented set of estimating equations that simultaneously satisfy the usual score/likelihood conditions from logistic regression, but also force the weighted covariate means to be equal between treated and control groups. In other words, balance constraints are baked directly into the equations used to solve for β, rather than only checked afterward. The method, therefore, avoids an iteration between model fitting and balance checking. Note that in the architecture of the CBPS model, we chose to target ATT (average treatment effect on the treated) rather than ATE since in general matching builds counterfactuals for only the treated group so all our estimates in this entire assignment have been estimating ATT and not ATE. Moreover, we use the "exact" method to ensure covariate means are matched exactly between groups.

```{r}
cbps.model <- CBPS(treatment.swang ~ ., data = rhc.selected[, c(xvars, "treatment.swang")],
                    ATT = 1, method = "exact") #Fit the CBPS model to the data
summary(cbps.model) #significance of coefficients etc. important for variable importance analysis later
pscore.cbps <- fitted(cbps.model)  # propensity scores
pscore.cbps.c <- pmin(pmax(pscore.cbps, 0.005), 0.995) #clipping as usual

brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.cbps, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.cbps[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "Propensity score distribution in treated vs untreated groups.", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap them nicely
hist(pscore.cbps[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlapping histogram


logits.cbps <- logit(pscore.cbps.c)

set.seed(42) #seed for reproducibility if multiple controls are at same distance, as explained before.
# Same matching and SMD calculation steps as fore the general GLM logistic regression model earlier.
psmatch.cbps <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logits.cbps,
                       replace = FALSE, Weight = 2, caliper = 0.2)
matched_cbps <- rhc.selected[unlist(psmatch.cbps[c("index.treated", "index.control")]), ]
matchedtab1_cbps <- CreateTableOne(vars = xvars, strata = "treatment.swang", data = matched_cbps, test = FALSE)
print(matchedtab1_cbps, smd = TRUE)
smd_values_matched_cbps <- ExtractSmd(matchedtab1_cbps)

sprintf("Matched pairs: %d (of %d treated)", length(psmatch.cbps$index.treated), sum(rhc.selected$treatment.swang))
sprintf("Max SMD: %.3f", max(smd_values_matched_cbps))
sprintf("SMD < 0.1: %d variables", sum(smd_values_matched_cbps < 0.1))
sprintf("0.1 <= SMD < 0.2: %d variables", sum(smd_values_matched_cbps >= 0.1 & smd_values_matched_cbps < 0.2))
sprintf("SMD > 0.2: %d variables", sum(smd_values_matched_cbps > 0.2))

```

#TODO write conclusion paragraph.



# Here we define our best performing model for each of the 2 finetuned methods and then continue with the rest of the analysis.
# Then, compare the number of matched pairs and balancing (through max SMD score, and number of variables below 0.1, between 0.1-0.2 and over 0.2 respecitvely). below is some code from before that can be used for this.
```{r}
sprintf("Matched pairs: Mahalanobis = %d, glm PS = %d, random forest PS = %d (of %d treated)",
        length(greedymatch$index.treated), length(psmatch$index.treated),
        length(psmatch.rf$index.treated), sum(rhc.selected$treatment.swang))
# All four stages side by side
smd.compare <- data.frame(before      = ExtractSmd(table1)[, 1],
                          mahalanobis = ExtractSmd(matchedtab1)[, 1],
                          ps.glm      = ExtractSmd(matchedtab1_ps)[, 1],
                          ps.rf       = ExtractSmd(matchedtab1_rf)[, 1])
print(round(smd.compare[order(-smd.compare$before), ], 3))
print("Variables with SMD > 0.1:"); print(colSums(smd.compare > 0.1))
print("Worst SMD:");                print(round(apply(smd.compare, 2, max), 3))
```


# This propensity predictive value performance analysis can be included when we compare the balancing versus the performance in how well it predicts propensity scores. So basically for each of the 4 methods, perform some type of propensity prediction analysis on held-out data / cross validation and on the full training data, to measure generalization vs overfitting, and compare & contrast these results to how well they balance the groups to explane the duality we discussed before between prediction performance and how well it balances. Code below is starter for RF model that calculates AUC (can maybe calculate AUC for all 4)? This is point 8. in the list below btw.

```{r}
set.seed(42)
#Here we redfine and fit our best performing RF model with node size = 10.
rf.model  <- randomForest(treatment.swang ~ ., data = rf.data, ntree = 500, nodesize = 10)
print(rf.model)
plot(rf.model)   # out-of-bag (held-out validation) error rate against number of trees: flat long before 500

oob.pscore.rf <- predict(rf.model, type = "prob")[, "1"]
auc.oob.rf <- auc(rf.data$treatment.swang, oob.pscore.rf)
train.pscore.rf <- predict(rf.model, newdata = rf.data, type = "prob")[, "1"]
auc.train.rf <- auc(rf.data$treatment.swang, train.pscore.rf)
sprintf("RF (nodesize=10): train AUC = %.3f, OOB AUC = %.3f", auc.train.rf, auc.oob.rf)
```
The final forest reports an out-of-bag error of 28.9%, against a no-information majority-vote baseline of 38.1% (the share of treated patients), so it has learned about who gets a catheter but is far from certain about it. The confusion matrix (see above) shows that this uncertainty is one-sided: 86% ((1-0.1368629) * 100) of untreated patients are classified correctly but only 46%  ((1-0.5366300) * 100) of treated ones. That is the majority class effect again, a patient is only labelled "treated" if more than half the trees vote for it, and with 62% of patients untreated the forest leans that way.#TODO: is that true? Howevre we dont use the 0.5 cut-off rather we use the vote fraction itself as the propensity score. A theme that echoes throughout this assignment is that perfect prediction would be bad, since it would mean treated and untreated patients do not overlap and cannot be matched. The error plot shows the out-of-bag error (black) settling by around 150 trees and staying flat to 500, with the two class errors (red for untreated, green for treated) doing the same, which confirms that 500 trees is plenty. 



# Here we need to do point 3. do the t-test/mcnemar test/ glm test to see if treatment explains outcome in each of the three models and compare
# just look at roughly what we did in the closed part and repeat that for each of the three models. It's already completed roughly for RF & NN in the code below, but not sure if we need the final effects dataframe to repeat the smd score reporting etc, since we did that before already. I think here just focus on odds ratio and p-values etc.

```{r}
print("===================================================")
y_trt.rf <- matched_rf$outcome.died[matched_rf$treatment.swang == 1]
y_con.rf <- matched_rf$outcome.died[matched_rf$treatment.swang == 0]

t.test(y_trt.rf - y_con.rf)                 # paired t-test, indicative only
table(y_trt.rf, y_con.rf)
mcnemar.test(table(y_trt.rf, y_con.rf))     # correct test for paired binary data

gmodel5 <- glm(outcome.died ~ treatment.swang, family = binomial, data = matched_rf)
summary(gmodel5)
sprintf("Odds ratio for treatment: %.4f", exp(coef(gmodel5)["treatment.swang"]))

# Doubly robust: adjust for the least well balanced variable
worst.rf <- rownames(smd_values_matched_rf)[which.max(smd_values_matched_rf)]
gmodel6 <- glm(as.formula(paste("outcome.died ~ treatment.swang +", worst.rf)), family = binomial, data = matched_rf)
summary(gmodel6)
sprintf("Odds ratio for treatment after adjusting for %s: %.4f", worst.rf, exp(coef(gmodel6)["treatment.swang"]))

effects <- data.frame(
  method      = c("Mahalanobis", "PS (glm)", "PS (random forest)"),
  pairs       = c(length(greedymatch$index.treated), length(psmatch$index.treated), length(psmatch.rf$index.treated)),
  mean.diff   = c(mean(y_trt - y_con), mean(y_trt.ps - y_con.ps), mean(y_trt.rf - y_con.rf)),
  OR          = exp(c(coef(gmodel1)["treatment.swang"], coef(gmodel3)["treatment.swang"], coef(gmodel5)["treatment.swang"])),
  OR.adjusted = exp(c(coef(gmodel2)["treatment.swang"], coef(gmodel4)["treatment.swang"], coef(gmodel6)["treatment.swang"])),
  smd.0.1.to.0.2 = colSums(smd.compare > 0.1 & smd.compare <= 0.2)[c("mahalanobis", "ps.glm", "ps.rf")], # number of variables with borderline imbalance (0.1 < SMD <= 0.2)
  smd.over.0.2 = colSums(smd.compare > 0.2)[c("mahalanobis", "ps.glm", "ps.rf")], # number of variables with clear imbalance (SMD > 0.2)
  worst.smd   = round(apply(smd.compare, 2, max)[c("mahalanobis", "ps.glm", "ps.rf")], 3))
print(effects, digits = 4, row.names = FALSE)




y_trt.nn <- matched_nn$outcome.died[matched_nn$treatment.swang == 1]
y_con.nn <- matched_nn$outcome.died[matched_nn$treatment.swang == 0]

t.test(y_trt.nn - y_con.nn)
mcnemar.test(table(y_trt.nn, y_con.nn))

gmodel7 <- glm(outcome.died ~ treatment.swang, family = binomial, data = matched_nn)
worst.nn <- rownames(smd_values_matched_nn)[which.max(smd_values_matched_nn)]
gmodel8 <- glm(as.formula(paste("outcome.died ~ treatment.swang +", worst.nn)), family = binomial, data = matched_nn)

sprintf("OR (unadjusted): %.4f", exp(coef(gmodel7)["treatment.swang"]))
sprintf("OR (adjusted for %s): %.4f", worst.nn, exp(coef(gmodel8)["treatment.swang"]))

effects <- rbind(effects, data.frame(
  method       = "PS (neural net)",
  pairs        = length(psmatch.nn$index.treated),
  mean.diff    = mean(y_trt.nn - y_con.nn),
  OR           = exp(coef(gmodel7)["treatment.swang"]),
  OR.adjusted  = exp(coef(gmodel8)["treatment.swang"]),
  smd.0.1.to.0.2 = sum(smd_values_matched_nn > 0.1 & smd_values_matched_nn <= 0.2),
  smd.over.0.2 = sum(smd_values_matched_nn > 0.2),
  worst.smd    = round(max(smd_values_matched_nn), 3)))
print(effects, digits = 4, row.names = FALSE)
print("===================================================")
```
So what does this mountain of numbers and graphs tell us? Let's break it down - The overlap is good, with the centre around 0.4, both groups covering 0.05 to about 0.85, with the forest not producing any scores near 1. The caliper keeps 1672 out of our 2184 treated patients, which is 62 fewer than the glm (TODO: #any conclusion about this?). The balance is similarly good, but not exceptional, no variable above SMD 0.2, 38 of 40 below 0.1, with only malighx (0.113) and pot1 (0.107) borderline (#TODO, compare to glm pscore, i think had everything below 0.1? should we compare mean smd score?). 
On the whole it beats the distance matching but lags a bit behind the glm. The effect points the same way as before but is smaller: mean difference 0.038 (t-test p = 0.021), McNemar p = 0.023 with 406 discordant pairs where the treated patient died against 343 where the control died, OR 1.18 (p = 0.022) versus 1.31 for the glm and 1.48 for Mahalanobis. Our preference is therefore the glm propensity score: it balanced every variable, kept more pairs, needed no tuning or clipping (#TODO: why?), did not move under adjustment, and is readable. The forest did not confirm our hypothesis that a threshold-based model would balance better, but it earns its place as a robustness check: a model with no additivity assumption, matched by the same protocol, still finds higher mortality among treated patients, so the RHC result does not depend on the form of the logistic model (#TODO love this). Mahalanobis matching is the clear loser, with the most residual imbalance (#TODO what do you mean with this) and the estimate that shifted most under adjustment #TODO and worst balance/SMD scores? or is that residual imbalance?.


# Here we need to do point 6. Do correlation analysis between propensity scores of the 4 methods. There is some code from before below but it is NOT complete yet. maybe some other statistical test to check correlation between the 4 different sets of propensity scores can be used, or 4x3=12 different cross tests, im not sure.
```{r}
sprintf("Correlation between glm and random forest propensity scores: %.3f", cor(pscore, oob.pscore.rf))
sprintf("Correlation with glm propensity scores: %.3f", cor(pscore, pscore.nn))
sprintf("Correlation with RF propensity scores: %.3f", cor(oob.pscore.rf, pscore.nn))
```
Finally, the forest's propensity scores correlate at 0.82 with those from the logistic regression. This means the two models largely agree on who was likely to be treated, which is reassuring, but they are not identical (TODO: inspect where they are different/same? like which patients the scores are more/less similar and if those patients share attributes?), so there is room for the forest's threshold-based view of the data to make a difference (#TODO what does that mean?). Whether it does is a matter of balance, so now we continue the analysis on the RF model with node size = 10, since that was the best performing model, exactly as we did on the glm model.


#Here still need to do 4. variable importance analysis. Like try to interpret coefficients. see below for little more explanation.

#Here we still need to do 5. caliper threshold analysis to try and find out if there is a certain threshold we prefer most, or if it depends on the model etc.


#Do Partial dependence plots and IPW on friday only if we have time.


# Add a final conclusion paragraph about if we prefer a certain model, or major conclusions we can draw.



#Okay to do gameplan so it's clear:
1. check what covariate balancing propensity score form imai& ratkovic 2014 is about. can we use it? CHECK, sounds great so let's use it!
2. Tune hyperparams based on the best balance(e.g. lowest max SMD/fewest variables over 0.1) since that is in the end what we care about, not the propensity prediction accuracy. 
3. do the t-test/mcnemar test/ glm test to see if treatment explains outcome in each of the three models and compare p-values etc. like in closed part roughly.

4. Variable importance analysis. For GLM (coefficient magnitudes/p-values), RF (important() and varImpPlot(), read claude about it.) (and if possible to some extent NN), do the two models agree on which variables matter most for predicting treatment? If they diverge substantially, that's an interesting, reportable finding (e.g., maybe the forest picks up a variable that's important only in combination with others — nonlinear/interaction effects — that the additive glm can't see)
5. Caliper threshold analysis, show bias/variance tradeoff

6. Do correlation analysis between propensity scores of the 4 methods? Not completely sure what this brings us but if we can make a nice story around it I think it's good to add. 

7. Partial dependence plots. Do this for GLM vs RF vs NN. How it's calculated: to see how the model's predicted P(treatment) depends on one variable (say, age), you take the entire dataset, artificially set every patient's age to a fixed value (e.g., age=50), get the forest's predicted probability for everyone under that assumption, and average those predictions. Repeat across a grid of age values (e.g., 20, 30, 40, ... 90) — the resulting curve shows how average predicted treatment probability moves as age moves, holding the "effect" of all other variables fixed at their observed values (each prediction still uses each patient's true values for everything else — only the one variable of interest is swept). In R: randomForest::partialPlot(rf.model, pred.data=rf.data, x.var="age").

What you learn: whether the relationship looks smooth/monotonic (similar to what a linear/logistic term would capture) or shows a sharp jump/plateau at some threshold (the "clinical cutoff" behavior your team hypothesized trees might capture that a glm's additive linear term cannot).

What I meant by "even though the matching/balance results didn't confirm it" / "negative result": your team's original hypothesis was: because RF thinks in thresholds like clinicians do, it should produce a better propensity score — and "better" was operationalized as better balance after matching (lower SMD, less extreme scores). That didn't pan out — the glm actually balanced better in your existing results. That's the "negative result" for the practical hypothesis. But that doesn't mean the underlying premise (RF captures threshold effects, glm doesn't) is false — it just means capturing threshold effects didn't translate into better matching performance in this case. A partial dependence plot could still show real threshold behavior in the forest even though it didn't help balance — which is actually a more interesting, nuanced finding than either "RF wins" or "RF loses": it separates "does the model behave the way we expected" from "does that behavior help for this specific downstream task," and reasoning explicitly about why a true modeling difference didn't yield a practical improvement is exactly the kind of critical thinking the TA is asking for.

8. Test GLM, RF & NN on train AUC vs cross val AUC, to see if propensity prediction accuracy/generalization correlates to best-balance. like if best balanced model turns out to have a (slightly) overfitted propensity prediction model because it squished propensity values more towards the middle to create overlap (and thus not the best propensity prediction model we could get) then that's an interesting observation and worth noting.

9. Use inverse probability weighting based on propensity scores as a second approach to matching.Can do if have time over. not necessary persé but would be a nice addition.



#References
Imai, K., & Ratkovic, M. (2014). Covariate balancing propensity score. Journal of the Royal Statistical Society Series B: Statistical Methodology, 76(1), 243-263.
