---
title: "Casual Inference Notebook"
output:
  html_document:
    df_print: paged
---

In this assignment we will exercise causal inference on the RHC (Right Heart Catheterization) dataset This dataset was used in Connors et al. (1996): The effectiveness of RHC in the initial care of critically ill patients. J American Medical Association 276:889-897. The dataset pertains to day 1 of hospitalization, i.e., the "treatment" variable swang1 is whether or not a patient received a RHC (also called the Swan-Ganz catheter) on the first day in which the patient qualified for a specific research study called SUPPORT.

What should you submit? You need to submit:
1. This Notebook (in .Rmd format) after filling in all your answers inside this notebook. You are required to do an action (write or fill-in code or write documentation) at the places marked by an arrow (=>). This arrow appears within the comments (and it will be preceeded by "#") or outside the comments in the text (without "#").
2. The Notebook in html format with all the results.

Please provide information about your group:
==> Student names: Kieran Carroll, Hugo Sokołowski-Katzer, Isa Oxenaar
==> Student numbers: 16503120, 1534349, xxx

#TODO: names, student numbers and disclaimer that all em dashes used were human written haha. LLm usage disclose anywhere?
*Note that the set.seed(42) is run in many cells, to ensure reproducibility of results even if cells are not run in order all the time.

DISCLAIMER: AI such as Claude Opus 5 was used in the making of this assignment. The predominant use was idea brainstorming, debugging as well as R code syntax assistance. All analyses, interpretations of results, and written answers are our own work. AI tools were not used to generate the substantive content of our responses or conclusions.

# Closed part (70% of the grade)

## Understand the dataset:
You can download the dataset from the course canvas. It is called "rhc.csv". Please Open https://hbiostat.org/data/repo/rhc.html and take a good look at the variables. Make sure you inspect the variables included in the dataset (they appear in a tabel below the text on the website). You can refer to the table whenever needed. 

Packages: install packages if needed
# ```{r}
# install.packages("tableone")
# install.packages("Matching")
# # Recommended: add "dplyr" if you have used it before to make efficient preprocessing.
# ```

## Load packages
```{r}
library(tableone) # This library allows us to calculate summary statistics about the variables including the SMD (standardized mean difference).
library(Matching) # This library will do the matching
```

## Read database
```{r}
# Download rhc dataset from CANVAS and read it
rhc <- read.csv("rhc.csv")
```

## Understanding the dataset  
Tip: Look at the data, use the "View" command on the dataset "rhc"
```{r}
print("===================================================")
# => how many rows does rhc contain? write a print command
# => What are the variable names? write a print command
number_rows <- nrow(rhc)
column_names <- names(rhc)
sprintf("There are %d rows in the rhc dataframe.", number_rows)
sprintf("The following are the column names in the rhc dataframe:")
print(column_names)
print("===================================================")
```

For the assignment the TREATMENT variable is swang1 and the OUTCOME variable is death.

## Preprocess
If you look at the values of the dichotomous variables you will see that they are factors (in other words, categorical variables). We would like to work with numeric variables in this exercise so we need to perform some conversions.

Let's start with the treatment and outcome variables. Create an "outcome.died" and "swang.treatment" variables with the numeric values 0 and 1. Later we will deal with the rest.
```{r}
rhc$outcome.died <- as.numeric(rhc$death=='Yes')  # Again you might want, instead, to do all preprocessing with dplyr if you prefer
rhc$treatment.swang <- as.numeric(rhc$swang1=='RHC')
```

### Check

Let's take a quick look into the dataset using these variables
```{r}
print("===================================================")
# => What is the mean of outcome.died in the whole data set?
mean_outcome_died <- mean(rhc$outcome.died)
sprintf("Mean of outcome.died in whole data set is: %.4f", mean_outcome_died)

# => How many subjects are treated?
num_treated <- sum(rhc$treatment.swang)
sprintf("There are %d subjects treated.", num_treated)

# =>  What is the mean mortality (outcome.died) in this treated group?
mean_mortality_treated <- mean(rhc$outcome.died[rhc$treatment.swang == 1])
sprintf("Mean mortality in treated group is: %.4f", mean_mortality_treated)

# => How many subjects are not treated (i.e. in the control group)?
num_not_treated <- sum(rhc$treatment.swang == 0)
sprintf("There are %d subjects not treated.", num_not_treated)

# => What is the mean mortality in this control group?
mean_mortality_not_treated <- mean(rhc$outcome.died[rhc$treatment.swang == 0])
sprintf("Mean mortality in not treated group is: %.4f", mean_mortality_not_treated)

print("===================================================")
```


The variable cat1 is a factor (categorical variable) containing the primary disease category for the patient.
```{r}
print("===================================================")
# => How many different categories are in Cat1? Tip: use the levels() function on this variable.
num_levels_cat1 <- length(levels(rhc$cat1))
class <-class(rhc$cat1)
sprintf("Cat1 is actually a %s class.", class)
# Convert to factor:
print("Converting to factor first, then counting number of different categories:")
cat1_factor = as.factor(rhc$cat1)
print("The different cat1 categories:")
print(levels(cat1_factor))
num_levels_cat1 <- length(levels(cat1_factor))
sprintf("Cat1 has %d different categories.", num_levels_cat1)
print("===================================================")
```

We would like to "translate" Cat1 into a set of variables each with the numeric values of 0/1. For example for the value of "ARF" we will create a variable "ARF" that will be 1 if rhc$cat1=='ARF'.

```{r}
rhc$ARF <- as.numeric(rhc$cat1=='ARF')
print("===================================================")
# => Do the same (as we did for ARF) also for all other values of cat1 but ignore the category "Missing".
rhc$CHF             <- as.numeric(rhc$cat1 == 'CHF')
rhc$Cirrhosis       <- as.numeric(rhc$cat1 == 'Cirrhosis')
rhc$Colon.Cancer    <- as.numeric(rhc$cat1 == 'Colon Cancer')
rhc$Coma            <- as.numeric(rhc$cat1 == 'Coma')
rhc$COPD            <- as.numeric(rhc$cat1 == 'COPD')
rhc$Lung.Cancer     <- as.numeric(rhc$cat1 == 'Lung Cancer')
rhc$MOSF.Malignancy <- as.numeric(rhc$cat1 == 'MOSF w/Malignancy')
rhc$MOSF.Sepsis     <- as.numeric(rhc$cat1 == 'MOSF w/Sepsis')
 print("===================================================")
```

## Create data frame with variables of interest

==> Create a data frame and call it rhc.small with the following variables of interest:
All the underlying categories of "cat1" (ARF etc) that you created before,   "cardiohx", "chfhx", "dementhx", "psychhx",  "chrpulhx", "renalhx", "liverhx",  "gibledhx", "malighx",  "immunhx",  "transhx",  "amihx", "age", "sex", "edu", "das2d3pc", "aps1", "scoma1", "meanbp1", wblc1", "hrt1",     "resp1", "temp1", "pafi1", "alb1", "hema1", "bili1", "crea1", "sod1", "pot1",     "paco21", "ph1", "wtkilo1", "dnr1", "resp", "card", "neuro", "gastr",  "renal", "meta", "hema", "seps", "trauma",  "ortho", "adld3p", "urin1", "treatment.swang", "outcome.died".

```{r}
# rhc.small <- => Create this data frame
rhc.small <- rhc[ , c("ARF", "CHF", "Cirrhosis", "Colon.Cancer", "Coma", "COPD",
                      "Lung.Cancer", "MOSF.Malignancy", "MOSF.Sepsis",
                      "cardiohx", "chfhx", "dementhx", "psychhx", "chrpulhx", "renalhx",
                      "liverhx", "gibledhx", "malighx", "immunhx", "transhx", "amihx",
                      "age", "sex", "edu", "das2d3pc",
                      "aps1", "scoma1", "meanbp1", "wblc1", "hrt1", "resp1", "temp1", "pafi1",
                      "alb1", "hema1", "bili1", "crea1", "sod1", "pot1", "paco21", "ph1", "wtkilo1",
                      "dnr1", "resp", "card", "neuro", "gastr", "renal", "meta", "hema", "seps",
                      "trauma", "ortho", "adld3p", "urin1",
                      "treatment.swang", "outcome.died")]
nrows <- nrow(rhc.small)
ncols <- ncol(rhc.small)
sprintf("rhc.small has %d rows and %d columns.", nrows, ncols)
```

Make sure you take a look at the definition of variables (on the URL given above) with unclear names to you. This data frame includes now all candidate variables from which we will select later for matching. However, before we jump into variable selection, let us continue data type conversion.

Check classes of these variables by using the str command
```{r}
str(rhc.small)
```

If you have "labelled" variables like labelled "int" and labelled "num" it means that the dataset was imported from SPSS. In such a case, simply change such variables into numeric variables. If you do not see such variables then skip this step.

```{r}
#To our understanding this is not the case with us, thus we skip this step.

#rhc.small$age <- as.numeric(rhc.small$age) # only if in your version of the data "age" was a "labelled" variable
print("===================================================")
# => Change all labelled variables that include numbers (integer or numeric) into numeric.
# print("===================================================")
```

For each  categorical variable with 2 values (such as sex) change it into a numeric variable by selecting one of the values (such as "female") to be 1 and the other zero.

```{r}
rhc.small$female <- as.numeric(rhc.small$sex=='Female')
print("===================================================")
# => Change any factor with two values into a 0/1 variable by selecting a value (such as "female" above) to be 1 and the other value will be zero.
rhc.small$dnr1   <- as.numeric(rhc.small$dnr1   == 'Yes')   
rhc.small$resp   <- as.numeric(rhc.small$resp   == 'Yes')  
rhc.small$card   <- as.numeric(rhc.small$card   == 'Yes')
rhc.small$neuro  <- as.numeric(rhc.small$neuro  == 'Yes')
rhc.small$gastr  <- as.numeric(rhc.small$gastr  == 'Yes')
rhc.small$renal  <- as.numeric(rhc.small$renal  == 'Yes')
rhc.small$meta   <- as.numeric(rhc.small$meta   == 'Yes') 
rhc.small$hema   <- as.numeric(rhc.small$hema   == 'Yes')
rhc.small$seps   <- as.numeric(rhc.small$seps   == 'Yes')
rhc.small$trauma <- as.numeric(rhc.small$trauma == 'Yes')
rhc.small$ortho  <- as.numeric(rhc.small$ortho  == 'Yes')

rhc.small$sex <- NULL #removing sex since it wasnt replaced in-place, without this it messed up the matching later on

#quick sanity check for whether all of the above worked
print(names(rhc.small)[sapply(rhc.small, is.factor)])
print("===================================================")
```
Before we start the analysis we need to be aware of the following issue: If you encounter a variable with missing values then you might run into a problem with the Match() function later on. For this exercise you can follow this advice:

1) If the proportion of missing is above 15% for a variable then just exclude the whole variable from the data frame and hence from the analysis, it is likely that the information it had still resides in the combination of the many other variables.

2) If you have less than 15% missing variables try to impute the values of the variable. In the extreme case you could apply multiple imputations and repeat the analysis on each imputation set and then pool the results. However for this exercise you can follow a simple strategy such as imputing the value with mean/median/mode. This would be a limitation but imputing missing values is not the focus of our exercise on causal inference.

```{r}

# => Inspect the proportion of missing values for each variable and follow the advice above.
print("===================================================")

#for every column (sapply) give True or False depending on whether it is na, then take the mean of that (treat True/False as 1/0)-> proportion of missing values for each variable.
missing.fraction <- sapply(rhc.small, function(column) mean(is.na(column)))
#print all variables that have missing values in descending order.
print("Variables that have missing values, sorted by descending proportion of missing values:")
print(sort(missing.fraction[missing.fraction > 0], decreasing = TRUE))

#take names of all the variables where the fraction is above the threshold (0.15) and drop those columns from the dataframe (keep all others). This got rid of sdld3p and urin1 which were actually the only two vars with missing values (see earlier printed list above), so the following imputation is redundant, nonetheless:
rhc.small <- rhc.small[ , !(names(rhc.small) %in% names(missing.fraction)[missing.fraction > 0.15])]
for (v in names(missing.fraction)[missing.fraction > 0 & missing.fraction <= 0.15]) {
  rhc.small[[v]][is.na(rhc.small[[v]])] <- median(rhc.small[[v]], na.rm = TRUE)
} #loops over variables that have a missing value proportion between 0 and 0.15 (excluding the former, including the latter), and impute the missing values for those variables as the median of the present values of that variable.
#We chose median as its generally more reliable and less skewed than average. And for data such as urin1 where the mode is 0, it offers a bit more information over just plugging the mode.


sprintf("There are %d missing values left in the rhc.small dataframe (should be 0)", sum(is.na(rhc.small)))
print("Sanity check that dimensions of rhc.small dataframe did not change:")
print(dim(rhc.small))

#Note: watch out because re-running this cell wont show the two dropped vars, since they no longer exist, do "run all" to see that
print("===================================================")
```
```{r}
# Check/see that all variables are numeric or int
str(rhc.small)
```
Now we are ready to start the analysis.

## Selecting variables to control for

According to the disjunctive cause criterion, we want to find variables (to control for) that are predictive of treatment and/or the outcome. In this assignment we will use the criterion to select all (pre-treatment) variables that are predictive of the outcome. This will include true confounders (that also affect the treatment, and risk factors that only affect the outcome). All our variables in rhc.small, except the outcome of course, are pre-treatment variables (none is measured AFTER the intervention) so they are all good candidates for the selection.

The truth is that the best way to select these variables is to use clinical knowledge. However, in this assignment we will do this in a data-driven and practical way. Specifically, we will operationalize "predictive of the outcome" as the following criterion: a variable is predictive of the outcome if its univariate association with the outcome is significant at the 0.1 level. Because our outcome is binary we can translate this criterion to: For every candidate variable in rhc.small fit a logistic regression model to predict outcome (outcome.died) using only the candidate variable as a predictor and retain the variable if the p-value of its association with the outcome is ≤ 0.1, and otherwise discard the variable. Note: Do not include the treatment in the selection. 

```{r}
print("===================================================")
# => Fit logistic regression models for each variable separately with the outcome and retain those with an association having a p-value of ≤ 0.1. Tip: you may want to automate the whole process!

#Why 0.1 and not 0.05? It's mainly because variable screening should have looser requirements than hypothesis testing because missing a possible confounder is much more harmfull (potentially increased bias significantly), compared to the additional cost of controlling for a variable that isn't actually a confounder.

#Make our candidates, but we first remove two variables the treatment and outcome, since checking if outcome affects outcome is pointless and we aren't experimenting on whether the treatment works, just looking for confounders, so ignoring those two:
candidates <- setdiff(names(rhc.small), c("treatment.swang", "outcome.died"))

p.values <- c()
#v is our placeholder, we iterate over each variable, plug it into the glm, the "family = binomial" flag sets it to logistic regression. summary(model)$coefficients gives the coefficient estimate, z-value, std error, and p-value. Index into it to extract p-value for variable coefficient.
for (v in candidates) {
  model <- glm(as.formula(paste("outcome.died ~", v)), family = binomial, data = rhc.small)
  p.values[v] <- summary(model)$coefficients[2, 4]
}
print(sort(p.values))


# => Put the names of the selected variables (without the treatment and outcome variables) in the variable called xvars by writing xvars <- c("ARF", ...) or do it automatically!
# => Print xvar

#filter out the variables with p-value <= 0.1 and put into xvars
xvars <- names(p.values)[p.values <= 0.1]

print("Variables with p-value <= 0.1:")
print(xvars)
sprintf("Number of fariables with p-value <= 0.1: %d", length(xvars))


# => Put these selected variables in a data.frame called rhc.selected that also includes the treatment and outcome variables to this data frame.
rhc.selected <- rhc.small[ , c(xvars, "treatment.swang", "outcome.died")] #Put everything back into the dataframe 
print("Dimensions of new dataframe with just selected variables:")
print(dim(rhc.selected))
# print("===================================================")
```


## Inspect balance before matching

## look at "table 1"
```{r}
print("===================================================")
#=> look at "table 1", complete the command by filling in the "..."
# CreateTableOne's strata argument tells it: "compute all these variable summaries (mean and stdev) separately for each level of this strata/grouping variable (treatment vs no treatment), then line them up side by side (plus the SMD between them). test=False ensures to ignore p-values since we want SMD and not significance tests that are sensitive to sample size, as discussed in the slides.
table1<- CreateTableOne(vars=xvars,strata="treatment.swang", data=rhc.selected, test=FALSE)
## include standardized mean difference (SMD)
print(table1, smd=TRUE)
## For which variables we do not have a good balance? Look at the slides for a guideline. 
# Expect to see several variables with high SMD score since the patients who got RHC and those who didn't were not randomized, so they likely differ on covariates like illnes, age, comborbidities etc. Here we quantify that: for each variable, is the distribution similar between the treated and not treated group? Then, after matching we can do this analysis again and see if it fixed some/all of the imbalances between groups, so that now the treatment does appear to be randomized across each covariate. 

# Following the slides:
smd_values = ExtractSmd(table1)
adequate_balance <- rownames(smd_values)[smd_values < 0.1]
not_too_alarming <- rownames(smd_values)[smd_values >= 0.1 & smd_values < 0.2]
imbalanced <- rownames(smd_values)[smd_values > 0.2]

sprintf("The following %d variables are adequately balanced between treatment and no treatment:", length(adequate_balance))
print(adequate_balance)

sprintf("The following %d variables are decently balanced between treatment and no treatment (on the border but not too alarming):", length(not_too_alarming))
print(not_too_alarming)

sprintf("The following %d variables are too imbalanced between treatment and no treatment:", length(imbalanced))
print(imbalanced)
print("===================================================")
```


## Greedy matching on Mahalanobis distance
Let us now do matching in the hope that the balance will get better so we can have a more valid inference.

```{r}
print("===================================================")

#Match() function has to randomly chose between equal best matches when replacement = False. Thus, set seed for reproducibility of results.
set.seed(42)


# We want to match the treated with the untreated subjects. Find 1 on 1 match.
# # => Complete the command
#greedymatch <- Match(Tr=,M=, X=rhc.selected[xvars], replace=FALSE) 
# read the documentation for Match() by typing ?Match.
# => What does replace=FALSE mean?
# => What is inside X in the command above?

# Tr is a vector indicating for each observation whether it belongs to the treatment or not treated group, M determines how many control instances we match to each treatment instance (here 1:1 matching) and X is a matrix (therefore added as.matrix, per the documentation) with all the variables (covariates) to match on (calculate the Magalanobis distance with). replace = False means matching without replacement, i.e. once an untreated subject has been used (matched), it can't be matched again against another treated subject. This forces each matched control to be unique, so the order of matching matters, and thus this typically increases bias relative to matching with replacement. If we match with replacement, the same "good" control can be matched against multiple treated subjects, which gives closer matches on average (lower bias), but increases variances, highlighting the inherent bias-variance tradeoff. Lastly, Weight = 2 specifies to use the true Mahalanobis distance as distance metric for matching, including covariance correction (to ensure to not double penalize difference along two correlated dimensions as if they were independent), rather than only standardizing by variance which is what the default inverse-variance-weighting does.
greedymatch <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = as.matrix(rhc.selected[xvars]), 
                      replace = FALSE, Weight = 2)

print("===================================================")
```

```{r}
print("===================================================")
greedymatch$index.treated # => What do you think this is?
greedymatch$index.control # => What do you think this is?
# greedymatch$index.treated is a vector containing for all treated observation in the matched dataset, theit observation (index/row) numbers in the original dataset (rhc.selected). greedymatch$index.control contains the same indices, but then for the control subjects in the matched dataset (those selected during matching). Notably, index.treated[i] and index.control[i] together identify the i-th matched pair: row index.treated[i] is a treated subject, and row index.control[i] is the untreated subject matched to them. So this means that these two vectors let you reconstruct exactly which original rows ended up in the matched dataset, and how they're paired together, e.g., rhc.selected[greedymatch$index.treated[1],] and rhc.selected[greedymatch$index.control[1], ] give you the first matched pair. Lastly, note that since M=1 and replace=FALSE, index.control should contain just as many values as index.treated and no repeated values (each control used at most once), while each treated subject is matched since no caliper was specified.
print("===================================================")
```


=> Are all treated subjects matched to a control one?

```{r}
matched <- rhc.selected[unlist(greedymatch[c("index.treated","index.control")]), ]
matched[1,]  # This is the first treated subject
print("===================================================")
matched[length(greedymatch$index.treated)+1,] # => What do you think this subject is?
# Because unlist(greedymatch[c("index.treated","index.control")]) concatenates all of index.treated first, then all of index.control, the resulting matched dataframe contains in rows 1 to length(index.treated) → all the treated subjects in matching order, and in rows length(index.treated)+1 to 2×length(index.treated) (since 1:1 matching) all the matched controls, in the same order as their treated counterparts. Therefore, matched[length(greedymatch$index.treated)+1,] is the control subject matched to the first treated subject (the one printed just above here).
print("===================================================")
```

=> Added analysis about number of matched patients.

```{r}
num_matched_treated = length(greedymatch$index.treated)
num_matched_control = length(greedymatch$index.control)
sprintf("Number of matched treated patients is %d, and the number of matched control patients is %d", num_matched_treated, num_matched_control)
print("Number of matched treated and control patients should be equal since 1:1 matching.")

num_not_matched_treated = num_treated - num_matched_treated
fraction_treated_matched = num_matched_treated/num_treated*100
num_not_matched_control = num_not_treated - num_matched_control
fraction_control_matched = num_matched_control/num_not_treated*100


sprintf("There are %d treated patients not matched, meaning that %.2f percent of treated patients got matched",num_not_matched_treated, fraction_treated_matched)
sprintf("There are %d control patients not matched, meaning that %.2f percent of control patients got matched",num_not_matched_control, fraction_control_matched)
print("===================================================")
```


Get table 1 for matched data with standardized differences
```{r}
print("===================================================")
# CreateTableOne's strata argument tells it: "compute all these variable summaries (mean and stdev) separately for each level of this strata/grouping variable (treatment vs no treatment), then line them up side by side (plus the SMD between them). test=False ensures to ignore p-values since we want SMD and not significance tests that are sensitive to sample size, as discussed in the slides.
matchedtab1 <- CreateTableOne(vars=xvars,strata="treatment.swang", data=matched, test=FALSE) # => Complete the command to create table 1 after matching
# => Print the balance with smd. 
print(matchedtab1, smd=TRUE)

  
# Expect to see less variables with high SMD scores after matching. 

# Following the slides:
smd_values_matched = ExtractSmd(matchedtab1)
adequate_balance_matched <- rownames(smd_values_matched)[smd_values_matched < 0.1]
not_too_alarming_matched <- rownames(smd_values_matched)[smd_values_matched >= 0.1 & smd_values_matched < 0.2]
imbalanced_matched <- rownames(smd_values_matched)[smd_values_matched > 0.2]

sprintf("The following %d variables are adequately balanced between treatment and no treatment:", length(adequate_balance_matched))
print(adequate_balance_matched)

sprintf("The following %d variables are decently balanced between treatment and no treatment (on the border but not too alarming):", length(not_too_alarming_matched))
print(not_too_alarming_matched)

sprintf("The following %d variables are too imbalanced between treatment and no treatment:", length(imbalanced_matched))
print(imbalanced_matched)

# => After matching how many "bad" variables with SMD > 0.2 are there?
print("After matching there are 7 bad variables with SMD > 0.2, compared to 12 before matching.")


# print("===================================================")
```

## Outcome analysis
```{r}
print("===================================================")
#n <- length(greedymatch$index.treated) #n = number of treated and control subjects in matched
#y_trt <- matched$outcome.died[1:n]# => Put here the vector of outcomes y (from matched$outcome.died) of those in the treatment group
#y_con <- matched$outcome.died[(n+1):(2*n)]# Put here the vector of outcomes y (from matched$outcome.died) of those in the control group

y_trt <-matched$outcome.died[matched$treatment.swang == 1]# => Put here the vector of outcomes y (from matched$outcome.died) of those in the treatment group
y_con <- matched$outcome.died[matched$treatment.swang == 0]# Put here the vector of outcomes y (from matched$outcome.died) of those in the control group
print("===================================================")
```

## Let's do a paired t-test to see if there are differences in death between the matched groups. This is not the best test because the data is binary but we want to just get a feeling of the differences.

```{r}
diffy <- y_trt-y_con # pairwise difference
t.test(diffy)
```

=> Look at the p-value. Is it below 0.05?
Yes, the p-value = 1.453e-11, so far below 0.05. 
=> What does this mean?
This means the mean difference in outcome (death) between treated and not treated groups (0.0911) is significantly different from 0. In other words, RHC patients-treated patients had significantly higher death rates than their matched controls. Note however that a t-test isn't ideal here since the binary data violates the normality assumption of a t-test.

## We want to use the McNemar test, which is the best choice for paired (matched) binary data
```{r}
table(y_trt,y_con)
# In this table you will see the 2x2 table:  
#         y_con
# ytreat     0     1
#       0    a     b
#       1    c     d

# Enter these a, b, c, and d numbers in the following McNemar test
# print("===================================================")
#mcnemar.test(matrix(c(),2,2)) # => Complete the command by filling the a, b, c, and d numbers in the correct order.
mcnemar.test(table(y_trt,y_con))
# => Is the p-value less than 0.05? What does this result mean?
#Yes, the p-value is 2.294e-11, much less than 0.05. This means the discordant pairs (b and c, where the outcome of treated and control disagreed) are imbalance, i.e., treatment status is associated with chance of death within matched pairs (evidence of treatment effect). Here, since c=538 is statistically significantly greater than b=339, there seems to be evidence that there is a higher chance of death among the treated patients within matched pairs, which is consistent with the t-test above.
# print("===================================================")
```

## The last way we want to analyse the outcome is with logistic regression
```{r}
gmodel1 <- glm(outcome.died ~ treatment.swang, family=binomial, data=matched)
print("===================================================")
# => Is the coefficient of treatment significant? you can use summary(gmodel1) or confint(gmodel1)
summary(gmodel1)
# In summary(gmodel1) we can see that the p-value for treatment is 4.34e-10, indicating that the coefficient of treatment is significant. 

# =>  What is the odds ratio that is associated with treatment? You can easily derive it from the coefficient (Google otherwise how to get the odds ratio from the coefficient)
sprintf("Coefficient of treatment is: %.5f.", coef(gmodel1)["treatment.swang"])
# The estimated treatment coefficient is 0.39461, thus the odds ratio associated with treatment is exp(0.39461) = 1.4838. This means that the odds of death increase by a factor of 1.4838 in the treatment group compared to the no treatment group.

# =>  If the coefficient is positive, what does that mean in terms of odds ratio?
# As explained above, if the coefficient is positive the odds ratio (exp(coefficient)) will be > 1, meaning that the treatment group has a higher chance of death compared to the not treated group.

# => If the coefficient is negative, what does that mean in terms of odds ratio?
# Meanwhile, if the coefficient is negative the odds ratio (exp(coefficient)) will be < 1, meaning that the treatment group has a lower chance of death compared to the not treated group.

# The great thing about modelling is that we can now again adjust for the variables with the worst SMD. This is called doubly-robust estimation. Adjust for the variable with the worst SMD that you encountered before. To adjust you just add the name of the variable in the logistic regression formula. => Complete the command
# aps1 is the variable with the worst SMD (0.501) after matching, thus we adjust for it in logistic regression:
gmodel2 <- glm(outcome.died ~ treatment.swang + aps1, family=binomial, data=matched)
summary(gmodel2)
# => What is the odds ratio for treatment now?
# In summary(gmodel2) we can see that the p-value for treatment is 2.57e-2, indicating that the coefficient of treatment is still significant (< 0.05). 
sprintf("Coefficient of treatment is: %.5f.", coef(gmodel2)["treatment.swang"])
# The estimated treatment coefficient is 0.19891, thus the odds ratio associated with treatment is exp(0.19891) = 1.2201. This means that the odds of death increase by a factor of 1.2201 in the treatment group compared to the no treatment group. Notably, this is a considerably smaller odds ratio than before adjusting for aps1 (note that the p-value is also a lot larger now (2.57e-2 vs 4.34e-10)).
# print("===================================================")
```


## Propensity score matching

Fit a propensity score model to predict treatment (not the outcome!). Use logistic regression with glm
```{r}
print("===================================================")
# now the outcome variable is treatment (treatment.swang), not outcome.died since we're modeling "who was likely to receive RHC, given their covariates". "~ ." includes all other variables as predictors (besides what we're modeling).
psmodel <- glm(treatment.swang ~ ., family = binomial, data = rhc.selected[, c(xvars, "treatment.swang")])# => Use the variables we found before in xvars in the first assignment to predict treatment

summary(psmodel)   # show coefficients etc
pscore<-psmodel$fitted.values # create propensity score

print("===================================================")
```

# Now it is time to check the positivity assumption
# let's check the overlap/positivity between the propensity scores for the two groups. We do that before the matching.
```{r}
print("===================================================")
#  => plot the density of the propensity scores of the treated and untreated. Tip: you might want to use par(mfrow=c(2,1)) to create 2 panels (2 rows, 1 column) and then plotting the first hist (histogram), with say, ylim=c(0,650) and col='darkblue', and the second histogram with ylim=c(650, 0) (note that we flipped the order) and col='darkblue'. This will plot the blue histogram as a normal histogram and the red one will be plotted upside down underneath the blue one so you can easily see the overlap/positivity.


brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "Propensity score distribution in treated vs untreated groups.", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

# => What do you think about the overlap/positivity? 
# Across most of the range of propensity scores there is substantial overlap/positivity (between 0.1 and 0.8), there exist comparable treated and untreated subjects to match at most points along the score. Only below 0.1 and between 0.8-0.9, there is little overlap/positivity, indicating a possible positivity violation in those small regions, which risks no valid comparison and thus few matches in those regions (since we will be using a caliper to ensure we avoid poor, distant matches). However, since this is real RHC data, we expected this decent but imperfect overlap/positivity in the tails, since, as we observed before, there simply are significantly more deaths (and thus higher propensity scores) in the treated group. In conclusion, the positivity assumption is okay here, but we note a possible limitation in the tails.


print("===================================================")
```

We want to do greedy matching on PS using Match with a caliper. Note that instead of the probability PS itself, we usually use the log odds (logit) of PS which is log(PS/(1-PS)) which is also equal to log(PS)-log(1-PS) . So let us first write a function (of one line) to calculate the logit.

```{r}
print("===================================================")
logit <- function(p) {log(p/(1-p))} # => Complete this function to calculate the log odds
print("===================================================")
```


Now we want to use Match() with rhc.selected$treatment.swang as treatement, with 1 matched control per treated subject, with X as the logit(pscore), with no replacement and with a caliper of 0.2

```{r}
print("===================================================")
#Match() function has to randomly chose between equal best matches when replacement = False. Note that this is much more likely here than before since now we are only matching on 1 variable (propensity) rather than all the variables in xvars. Thus, set seed for reproducibility of results.
set.seed(42)

# Compute all logits
logits <- sapply(pscore, logit)

#psmatch <- Match(Tr=mydata$treatment,...) # => complete the command

psmatch <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logits, 
                      replace = FALSE, Weight = 2, caliper = 0.2)
print("===================================================")
matched_ps <- rhc.selected[unlist(psmatch[c("index.treated","index.control")]), ]
```


## Get standardized differences. Note: xvars is the same as in the first assignment.

```{r}
print("===================================================")
matchedtab1_ps <- CreateTableOne(vars=xvars, strata ="treatment.swang", 
                            data=matched_ps, test = FALSE) # => Complete the command
# => print the matching table showing the SMD
print(matchedtab1_ps, smd=TRUE)

# Expect to see less variables with high SMD scores after matching. 

# Following the slides:
smd_values_matched_ps = ExtractSmd(matchedtab1_ps)
adequate_balance_matched_ps <- rownames(smd_values_matched_ps)[smd_values_matched_ps < 0.1]
not_too_alarming_matched_ps <- rownames(smd_values_matched_ps)[smd_values_matched_ps >= 0.1 & smd_values_matched_ps < 0.2]
imbalanced_matched_ps <- rownames(smd_values_matched_ps)[smd_values_matched_ps > 0.2]


sprintf("Max SMD: %.3f", max(smd_values_matched_ps))

sprintf("The following %d variables are adequately balanced between treatment and no treatment:", length(adequate_balance_matched_ps))
print(adequate_balance_matched_ps)

sprintf("The following %d variables are decently balanced between treatment and no treatment (on the border but not too alarming):", length(not_too_alarming_matched_ps))
print(not_too_alarming_matched_ps)

sprintf("The following %d variables are too imbalanced between treatment and no treatment:", length(imbalanced_matched_ps))
print(imbalanced_matched_ps)

# => Are there any variables with bad balance now?
print("After matching there are 0 bad variables with SMD > 0.2 (even 0 with SMD > 0.1), compared to 7 after covariate matching and 12 before matching.")
print("===================================================")
```

## Outcome analysis after propensity score matching
```{r}
print("===================================================")
y_trt.ps <- matched_ps$outcome.died[matched_ps$treatment.swang == 1] # => Put here the vector of outcomes y (from matched$outcome.died) of those in the treatment group
y_con.ps <- matched_ps$outcome.died[matched_ps$treatment.swang == 0]# Put here the vector of outcomes y (from matched$outcome.died) of those in the control group
print("===================================================")
```


Perform a t-test
```{r}
# Perform paired t-test. Just to get a feeling of the result
diffy.ps <- y_trt.ps - y_con.ps # pairwise difference
t.test(diffy.ps)
print("===================================================")
# => Is the test significant?
#Yes, the p-value = 9.767e-06, so well below 0.05. 
# This means the mean difference in outcome (death) between treated and not treated groups (0.07151096) is significantly different from 0. In other words, RHC patients-treated patients had significantly higher death rates than their matched controls. Note however that a t-test isn't ideal here since the binary data violates the normality assumption of a t-test.
print("===================================================")
```


Use the McNemar test as before. Enter the a, b, c, and d numbers you get from the 2x2 table in the McNemar test
```{r}
print("===================================================")

table(y_trt.ps,y_con.ps)

#mcnemar.test(matrix(c(...),2,2)) # => Complete the command
# => Is the p-value less than 0.05? What does this result mean?
# => Compare the difference in the means before and after matching.


# In this table you will see the 2x2 table:  
#         y_con.p
# ytreat.p     0     1
#       0    a     b
#       1    c     d

mcnemar.test(table(y_trt.ps,y_con.ps))
# => Is the p-value less than 0.05? What does this result mean?
#Yes, the p-value is 1.208e-05, a lot less than 0.05. This means the discordant pairs (b and c, where the outcome of treated and control disagreed) are imbalance, i.e., treatment status is associated with chance of death within matched pairs (evidence of treatment effect). Here, since c=457 is statistically significantly greater than b=333, there seems to be evidence that there is a higher chance of death among the treated patients within matched pairs, which is consistent with the t-test above.
# => Compare the difference in the means before and after matching.
# Difference in means y_treat - y_con after matching on propensity is 0.07151096  (see t-test right above).
# Difference in means y_treat - y_con after matching on covariates is 0.09111722  (see first t-test).
# Difference in means y_treat - y_con before matching is 0.6804 - 0.6297 (from original descriptive data analysis at the top) = 0.0507. 
# Since propensity matching achieved better covariate balance (lower SMDs), it more successfully removed confounding on the measured covariates, so it's the more trustworthy estimate of the two. That the raw difference in means estimate happens to resemble the propensity-matched estimate is likely coincidental sample-specific behavior, and not proof that confounding was absent. The SMD tables tell us that propensity matching demonstrably balanced the covariates better, which is why its estimate is more trustworthy. The Mahalanobis number being even slightly higher is more likely an artifact of incomplete balancing (and/or a different matched subsample) than a genuine midpoint in some meaningful confounding-removal trajectory. Anyways, the difference in means after balancing the matches well in the propensity-based approach, indicate that after adjusting for those possible confounders, there is still a difference in means between the treated and not treated groups, which is in line with the significance found in the t-test, McNemar test and logistic regression below.

print("===================================================")
```

Use logistic regression
```{r}
gmodel3 <- glm(outcome.died ~ treatment.swang, family=binomial, data=matched_ps)
print("===================================================")
# => What is the odds ratio of treatment?
summary(gmodel3)
# In summary(gmodel3) we can see that the p-value for treatment is 0.000189, indicating that the coefficient of treatment is significant. 
sprintf("Coefficient of treatment is: %.5f.", coef(gmodel3)["treatment.swang"])
# The estimated treatment coefficient is 0.26655, thus the odds ratio associated with treatment is exp(0.26655) = 1.3055. This means that the odds of death increase by a factor of 1.3055 in the treatment group compared to the no treatment group.


# => Now correct also for the least well balanced variable
# scoma1 is the variable with the worst SMD (0.037) after matching, thus we adjust for it in logistic regression:
gmodel4 <- glm(outcome.died ~ treatment.swang + scoma1, family=binomial, data=matched_ps)
# => What is the odds ratio of treatment now?
summary(gmodel4)
# In summary(gmodel4) we can see that the p-value for treatment is 0.000163, indicating that the coefficient of treatment is still significant (<0.05). 
sprintf("Coefficient of treatment is: %.5f.", coef(gmodel4)["treatment.swang"])
# The estimated treatment coefficient is 0.27114, thus the odds ratio associated with treatment is exp(0.27114) = 1.3115 This means that the odds of death increase by a factor of 1.3115 in the treatment group compared to the no treatment group. Notably, this is a slightly larger odds ratio than before adjusting for scoma1, highlighting the fact that even the variable with the highest SMD value was still not significantly confounding the results of gmodel3.

print("===================================================")
```


# Open part (30% of the grade)
Instead of the glm to calculate the propensity score use any machine learning technique you like such as a regression tree (using rpart), or random forests (randomForest) or neural networks (neuralnet) or another formalism in R to create the propensity score model and to predict the probabilities on such a model. Note that you might need to tweak the model parameters in order to improve it. Compare your results to those you obtained by matching (via the Mahalanobis distance and the propensity score with logistic regression). Which approach do you prefer, and why?

```{r}
#Installing and loading required packages. Can comment installation lines out after ran once.
# install.packages("randomForest")
# install.packages("caret")
# install.packages("data.table") # reinstall into the user library: the system copy was built for an older R and fails to load (undefined symbol SETLENGTH), which breaks library(caret)
# install.packages("pROC")
# install.packages("CBPS")
# install.packages("ggplot2")
# install.packages("nnet")
# install.packages("corrplot")
# install.packages("survey")
library(CBPS)
library(randomForest)
library(pROC)
library(caret)
library(ggplot2)
library(nnet)
library(corrplot)
library(survey)
```


So far we have compared the Mahalanobis distance and propensity score, with the latter coming in as the strong favourite due to its more comprehensive nature - in the end encoding within it more information than the euclidian distance matching could. However we still have the case of machine learning models. As a group we have decided to try multiple approaches and compare our findings and results between them, looking at different performance metrics and characteristics. 


We decided to compare a random forest (RF), a neural network, and covariate balanced propensity score (CBPS) against the standard glm because each represents a genuinely different strategy for estimating the propensity score, and together they probe different weaknesses of the baseline logistic model. The glm assumes a strictly additive relationship, that is linear in the log odds, between covariates and treatment — a strong, potentially wrong assumption we take simply based on faith. Random forest relaxes this assumption by learning through hierarchical thresholds rather than smooth linear terms, closer to how clinical decisions are actually made (a lab value crossing a cutoff, not a proportional risk score), and is able to capture nonlinearities and interactions that the standard glm isn't able to express. On the otherhand, the neural network gives a different kind of flexibility again; it is a smooth, continuous function approximator that can, in principle, capture arbitrary nonlinear relationships (including interactions) without the discrete splits a tree is restricted to. However, as a result, it is far less interpretable and more sensitive to tuning and preprocessing. Lastly, CBPS (Imai & Ratkovic, 2014) is different from the other two: rather than adding flexibility to the functional form, it keeps the same linear logistic form as the glm but changes directly what the model is optimized for (adds constraints to the loss). This explicitly targets covariate balance in its fitting procedure rather than optimizing predictive likelihood alone. Testing all four side by side lets us separate distinct questions that are usually mixed up: does a more flexible/powerful functional form produce a better propensity score (RF, NN vs. glm), if so, does that produce better balanced covariate groups, and does optimizing directly for the right objective matter more than model flexibility at all (CBPS vs. glm)?

As a first optimization step and comparison between models, we will be finetuning some of the hyperparameters of the RF and NN models to optimize for covariate balance. Very important here is to emphasize the difference between propensity score predictive power and ability to create well-balanced covariate groups. The actual downstream of the propensity scores is to use them to create well balanced groups, not to perfectly classify each patient as treated/not treated. These two goals are related but not the same, and can actively conflict (e.g. a model that predicts treatment too well tends to push scores toward 0 and 1 (near-certain predictions), which destroys overlap/positivity). Maybe a model that is (slightly) overfitted balances better because it moves the propensity scores away from 0/1 and more towards the middle values which we results in more overlap/positivity and better positivity. Therefore, we optimize hyperparameters based on how well they balance the covariates by for each combination of hyperparameters look at the max SMD score, and the number of variables that have below 0.1, between 0.1-0.2 and over 0.2 SMD scores respectively.Note that for the baseline GLM and CBPS models there is no hyperparameter to do, since they essentially have fixed capacity once the formula/functional form is fixed. We will, however, investigate the addition of interaction terms in those two models. 

First up is Random Forest. It was primarily chosen due to its simplicity; it is often the classic first choice when ML is in the conversation, it may not have stellar results compared to other methods but it is steady, reliable and relatively interpretable. An interesting concept that came to mind (only a theory at this point, yet to be proven or disproven later in this chapter) is that of continuous vs discrete values. You see our previous two metrics, particularly PS is a continuous value, which works perfectly fine, but in the medical field our decisions are rarely motivated by proportions, but rather by thresholds. A patient is not "62% hypertensive", but rather a certain blood pressure threshold triggers treatment. Clinical reasoning can be observed as a sequence of cutoff points, which coincidentally is the precise mechanism a decision tree, and by extension - Random Forest, operates with.
Whether this structural resemblance translates into better estimates, is what we set out to test in this next section.

```{r}
rf.data <- rhc.selected[, c(xvars, "treatment.swang")]
rf.data$treatment.swang <- as.factor(rf.data$treatment.swang)
set.seed(42) #standard seed initialization for reproducibility

```

With the base data set up, its time for our hyperparameters and model itself

```{r}
rf.model <- randomForest(treatment.swang ~ ., data = rf.data, ntree = 500, nodesize = 25)
print(rf.model)
plot(rf.model)

```
The final forest reports an out-of-bag error of 28.9%, against a no-information majority-vote baseline of 38.1% (the share of treated patients), so it has learned about who gets a catheter but is far from certain about it. The confusion matrix (see above) shows that this uncertainty is one-sided: 86% ((1-0.1368629) * 100) of untreated patients are classified correctly but only 46%  ((1-0.5366300) * 100) of treated ones. That is the majority class effect again, a patient is only labelled "treated" if more than half the trees vote for it, and with 62% of patients the tree is biased that way. However, note that we don't use the 0.5 cut-off, rather we use the vote fraction itself as the propensity score. Again, it is important to state that perfect prediction would be bad, since it would mean treated and untreated patients do not overlap/positivity and cannot be matched. The error plot shows the out-of-bag error (black) settling by around 150 trees and staying flat to 500, with the two class errors (red for untreated, green for treated) doing the same, which confirms that 500 trees is plenty. We started with the default of 500 trees in the RF. Adding more trees never hurts the result, but adds compute time, and also by plotting rf.model we can see that the probabilities even out and are stable well before the 500 mark, suggesting that upping the number is not necessary. As such, from here on we only use 300 trees.

The two main dials we have are node size and the number of variables randomly considered as candidates at each split. Node size is the minimum number of patients allowed in a leaf. Its default is 1, meaning a tree keeps splitting until every leaf is pure, so in the limit each leaf is essentially one patient and the tree has memorized the training data. Our starting idea was that this would be bad for a propensity score: a leaf that has memorized a single patient can only ever say "treated" or "not treated", so we expected the scores to pile up at 0 and 1, which would give infinite logits and destroy the overlap/positivity we need for matching. Forcing larger leaves seemed like the natural fix, since a leaf of 25 patients has to base its vote on a mix of treated and untreated cases rather than on a single one, and we expected that to produce smoother scores and better overlap/positivity. We therefore chose node size = 25 as a starting point, large enough to stop memorization but small enough that the trees can still tell patients apart. Secondly, the default for the number of variables randomly considered as candidates at each split is sqrt(number of predictors), so ~ sqrt(40) which is roughly 6 here. A lower number of possible variables increases randomness and thus decorrelation between trees, which makes them less prone to overfitting but individually have weaker predictive power. Since it brings the propensity scores closer, we suspect better overlap/positivity and consequently better matching, and therefore expect lower mtry values (4/6) to perform better balancing. A larger number of possible variables make the trees more similar to each other and more powerful, but also more prone to overfitting. To test this reasoning and find out which values are optimal for balancing, we test it. 


For each of several node size values and number of randomly selected candidate variables we fit a forest, take its propensity scores, and run exactly the same matching and balance check we used for the glm propensity score, so that node size is judged by the thing we actually care about: overlap/positivity in the histograms and balance after matching. Important to note in this analysis is that initially the overlap/positivity checks were done strictly before matching. However, we believe it is better to analyze these results together, since it is not always true that overlap/positivity that seems better visually, also actually gives better balancing. On the other hand, it is also not always true that better SMD scores indicate strictly better balancing, since their might be little overlap/positivity resulting a small number of matched pairs, which is undesirable. Note that we use the number of matched pairs as a proxy for the overlap/positivity of the propensity distributions since less/more overlap/positivity results in fewer/more matched pairs when using a the standard 0.2*STD caliper, respectively. Therefore, we analyzed and reported the SMD scores together with the number of matched pairs, and also visually inspected the overlap/positivity graphs (of which a subset was printed for succinctness).


```{r}
nodesizes <- c(1, 5, 10, 25, 50, 100)
mtrys <- c(2, 3, 4, 6, 10, 15)  # default for classification is floor(sqrt(n_vars) ~= 6 here, so spread around it but not useful to go lower than 3 since than you get almost entirely random splits in the trees.

ps.by.combo <- list()
rf.comparison <- data.frame()

for (ns in nodesizes) {
  for (mt.val in mtrys) {
    set.seed(42) #Makes the forest's randomness (e.g. bootstrap sampling and feature selection at splits) reproducible
    m <- randomForest(treatment.swang ~ ., data = rf.data, ntree = 300, nodesize = ns, mtry = mt.val) #Fits forest for current nodesize/mtry combo
    ps <- predict(m, type = "prob")[, "1"] #Propensity score = fraction of trees voting "treated"
    
    key <- paste0("ns", ns, "_mtry", mt.val) #Unique key per combination, e.g. "ns10_mtry6"
    ps.by.combo[[key]] <- ps #Store propensity scores keyed by combo

    # Scores of exactly 0 or 1 have an infinite logit, so we push them up to 0.005 or down to 0.995 to not get an error
    ps.c <- pmin(pmax(ps, 0.005), 0.995)
    # 1:1 greedy matching with Mahalanobis distance on the logit transformed propensity score predicted by the RF, with no replacement and a caliper of      0.2 as before so that we can accurately compare the propensity score calculation method (controlling for other hyperparameters). 
    mt <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logit(ps.c),
                replace = FALSE, Weight = 2, caliper = 0.2)
    md <- rhc.selected[unlist(mt[c("index.treated", "index.control")]), ] #Extract matched data as before
    smd <- ExtractSmd(CreateTableOne(vars = xvars, strata = "treatment.swang", data = md, test = FALSE)) #Calculate SMD scores to check balancing

    # Comparison table; add 1 row of info per hyperparam combo
    rf.comparison <- rbind(rf.comparison, data.frame(
      nodesize       = ns,
      mtry           = mt.val,
      oob.error      = round(tail(m$err.rate[, "OOB"], 1), 3), #Out-of-bound error rate, i.e. held-out validation error for each full RF of 500 trees (similary to classical cross validation). Tells us how accurately predictive model is, but note that we don't want model to predict propensity scores perfectly as 0/1 since then there is bad overlap/positivity in scores. 
      extreme.scores = sum(ps < 0.01 | ps > 0.99), # patients with almost certain predictions, don't want this high since leads to unbalanced dataset (bad for positivity)
      matched.pairs  = length(mt$index.treated), # how many subjects succesfully found a caliper satisfying match (fewer = worse overlap/positivity)
      max.smd        = round(max(smd), 3), #maximum found smd (most unbalanced variable)
      worst.var = rownames(smd)[which.max(smd)], #variable with max (worst) smd score
      n.smd.under.0.1 = sum(smd < 0.1), #number of variables well-balanced
      n.smd.0.1.0.2  = sum(smd >= 0.1 & smd <= 0.2), #number of variables in danger zone
      n.smd.over.0.2 = sum(smd > 0.2))) #number of variables not well-balanced
  }
}
# sorted so the best-balancing combos appear first
print("Best balanced variables by max SMD value:")
print(rf.comparison[order(rf.comparison$max.smd), ]) 
print("Best balanced variables by number of variables well-balanced:")
print(rf.comparison[order(-rf.comparison$n.smd.under.0.1), ])

# heatmap of the balance metric to show balance results: number of variables well balanced (SMD < 0.1) and max SMD value.
ggplot(rf.comparison, aes(x = factor(nodesize), y = factor(mtry), fill = max.smd)) +
  geom_tile() +
  geom_text(aes(label = max.smd), size = 3) +
  scale_fill_gradient(low = "darkgreen", high = "darkred") +
  labs(x = "nodesize", y = "mtry", fill = "max SMD",
       title = "RF balance (max SMD) across nodesize x mtry grid") +
  theme_minimal()


ggplot(rf.comparison, aes(x = factor(nodesize), y = factor(mtry), fill = n.smd.under.0.1)) +
  geom_tile() +
  geom_text(aes(label = n.smd.under.0.1), size = 3) +
  scale_fill_gradient(low = "darkred", high = "darkgreen") +
  labs(x = "nodesize", y = "mtry", fill = "# SMD < 0.1",
       title = "RF balance (variables with SMD < 0.1) across nodesize x mtry grid") +
  theme_minimal()
```

These heatmaps tell a different story from the one we expected. For the node size, small values (1, 5 and 10) look best: the lower maximum smd values and more variables well-balanced (smd score < 0.1). Larger values make things worse, at a node size of 50 and 100 the max smd scores rise and a lot more variables are still unbalanced (up to 7 of the 40 variables have smd score > 0.1). The reason is that each tree votes for the majority class of its leaf, and with leaves of 100 patients the leaves are significantly biased to predicting majority-untreated since 62% of all patients in the entire dataset untreated, so the majority of trees vote "not treated" for most patients, bringing the propensity scores down and worsening matching as a result. Our worry about the forests remembering patients applied to a single tree, but the propensity score is the average vote of 300 trees grown on different bootstrap samples and different random subsets of features, and that averaging combined with the randomness already smooths the scores sufficiently. This hypothesis is confirmed in the first row of the figure below, where we indeed see that the overlap/positivity of propensity scores of patients is significantly worse at larger node sizes (and the default mtry = 6) due to the collapse into a spike near 0 by the majority vote phenomenon explained above. This consequently led to worse matches and thus lower smd scores as just discussed.
For mtry (the number of variables randomly selected as candidates at each split in a tree), we see that the best smd scores are obtained at the value of 3. Note that the default is sqrt(number of variables) = sqrt(40), which is roughly 6, so clearly in this situation the model prefers some added randomness (since fewer possible random candidate variables makes the splits less optimal and more decorrelated). A plausible explanation for this observation is that if the number of random candidate variables were higher, the splits would become more optimal, and thus the propensity scores would be predicted more accurately, possibly driving the treated patient's scores toward 1 and the control patient's scores to 0, reducing overlap/positivity and decreasing matching quality and thus balancing. Therefore, more randomness gives less optimal predictions and thus more overlap/positivity between propensity scores and consequent better balancing (better SMD metrics). Note that the overlap/positivity graphs show results in line with this. In row two of the figure below we see that higher mtry have slightly more spread out propensity values (contributing to worse matches and thus smd scores/balancing). 

Thus, looking at the table and heatmaps the combinations of nodesize 10 and mtry values 3 gives the best balance (0 variables with SMD above 0.1, max SMD 0.094), and the number of matched pairs is relatively good as well (roughly 1660) so that is what we use for the final random forest model. In the graph below we can observe that this configuration still leeds to good balancing; one hump per group, both spread across roughly 0 to 0.9, and wide overlap/positivity in the middle. This turned out to be practially the best overlap/positivity present in any of the configurations, and have the best balanced groups. Our starting guess of node size 25 and mtry 4/6 was close, but not the best.

```{r}
par(mfrow = c(2, 2))

pscore.rf <- pmin(pmax(ps.by.combo[["ns100_mtry6"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.rf, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.rf[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "ns = 100, mtry = 6", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.rf[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.rf <- pmin(pmax(ps.by.combo[["ns10_mtry6"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.rf, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.rf[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "ns = 10, mtry = 6", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.rf[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram


pscore.rf <- pmin(pmax(ps.by.combo[["ns10_mtry3"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.rf, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.rf[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "ns = 10, mtry = 3", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.rf[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.rf <- pmin(pmax(ps.by.combo[["ns10_mtry15"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.rf, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.rf[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "ns = 10, mtry = 15", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.rf[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram


par(mfrow = c(1, 1))
```


Next up is the neural network. As discussed before, random forests reason through discrete, threshold-based splits, while a a neural network offers the opposite kind of flexibility: a smooth, continuous function of the covariates, fit through gradient-based optimization rather than recursive partitioning. Since this let's us capture nonlinear relationships and interactions between covariates without being restricted to axis-aligned splits, we expect this to be able to optimize any goal relatively well. However, since this is a very powerful model, we need to ensure careful preprocessing (standardizing  numerically encoded inputs) and use regularization to avoid overfitting on a such a tiny dataset. As with random forest, we tune the most important hyperparameters: capacity (hidden layer size) and regularization strength (weight decay). However, again, following our own critique of predictive-accuracy-driven tuning, we select them based on the balance they produce after matching, not on how well they predict treatment assignment. We expect the model configuration that predicts the propensity scores well, but not too well, to have the best results since this will create seperated but still overlapping propensity scores, allowing for good balancing of the groups.


```{r}
# Preprocessing  (RF didn't need this)
nn.data <- rhc.selected[, c(xvars, "treatment.swang")]
nn.data$treatment.swang <- factor(nn.data$treatment.swang, labels = c("no", "yes"))  # nnet-friendly labels

# model.matrix handles factor -> one-hot expansion automatically and -1 drops intercept column
X.raw <- model.matrix(~ . - 1, data = nn.data[, xvars])

# Standardize (mean 0, sd 1): essential for NN training stability, unlike RF/glm
preproc <- preProcess(X.raw, method = c("center", "scale"))
X.scaled <- predict(preproc, X.raw)

nn.df <- data.frame(X.scaled, treatment.swang = nn.data$treatment.swang)

grid <- expand.grid(size  = c(1, 2, 4, 6, 8, 12),            # number of hidden units
                     decay = c(0, 0.001, 0.01, 0.1, 0.5))  # L2 regularization strength

ps.by.combo.nn <- list()
nn.comparison <- data.frame()

for (i in 1:nrow(grid)) { #loop over grid of hyperparams
  sz <- grid$size[i]; dc <- grid$decay[i]
  set.seed(42) #seed for reproducibility of random intial weights etc.
  # Fit an NN for a max of 300 epochs on this combination of hidden size / decay values. trace = FALSE removes printing at each epoch. linout = FALSE makes it output probability (through sigmoid) rather than raw logit.
  m <- nnet(treatment.swang ~ ., data = nn.df, size = sz, decay = dc,
            trace = FALSE, maxit = 300, linout = FALSE) 
  ps <- predict(m, newdata = nn.df, type = "raw")[, 1] #propensity scores for each patient. Note that we just make it predict on full training data and we don't care so much about overfitting, since, as explained before, we just watch propensity scores that balance well, we don't care about if they over or underfit the propensity prediction a bit in the end. 

  key <- paste0("size", sz, "_decay", dc)
  ps.by.combo.nn[[key]] <- ps

  # Scores of exactly 0 or 1 have an infinite logit, so we clip them
  ps.c <- pmin(pmax(ps, 0.005), 0.995)
  #Identical matching and SMD calculation steps as before
  mt <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logit(ps.c),
              replace = FALSE, Weight = 2, caliper = 0.2)
  md <- rhc.selected[unlist(mt[c("index.treated", "index.control")]), ]
  smd <- ExtractSmd(CreateTableOne(vars = xvars, strata = "treatment.swang", data = md, test = FALSE))

  # record hyperparam combination with same metrics as before
  nn.comparison <- rbind(nn.comparison, data.frame(
    size            = sz,
    decay           = dc,
    extreme.scores  = sum(ps < 0.01 | ps > 0.99),
    matched.pairs   = length(mt$index.treated),
    max.smd         = round(max(smd), 3),
    worst.var = rownames(smd)[which.max(smd)], 
    n.smd.under.0.1 = sum(smd < 0.1), 
    n.smd.0.1.0.2  = sum(smd >= 0.1 & smd <= 0.2), 
    n.smd.over.0.2 = sum(smd > 0.2)))
}

print("Best balanced combos by max SMD value:")
print(nn.comparison[order(nn.comparison$max.smd), ])   #ascending: lowest (best) max SMD first

print("Best balanced combos by number of variables well-balanced:")
print(nn.comparison[order(-nn.comparison$n.smd.under.0.1), ])  #descending: most (best) well-balanced vars first

# Again, heatmaps showing the balancing of different hyperparam combinations.
ggplot(nn.comparison, aes(x = factor(size), y = factor(decay), fill = max.smd)) +
  geom_tile() +
  geom_text(aes(label = max.smd), size = 3) +
  scale_fill_gradient(low = "darkgreen", high = "darkred") +
  labs(x = "size", y = "decay", fill = "max SMD",
       title = "NN balance (max SMD) across size x decay grid") +
  theme_minimal()

ggplot(nn.comparison, aes(x = factor(size), y = factor(decay), fill = n.smd.under.0.1)) +
  geom_tile() +
  geom_text(aes(label = n.smd.under.0.1), size = 3) +
  scale_fill_gradient(low = "darkred", high = "darkgreen") +
  labs(x = "size", y = "decay", fill = "# SMD < 0.1",
       title = "NN balance (variables with SMD < 0.1) across size x decay grid") +
  theme_minimal()
```
Interestingly, what stands out at first hand are the amazing smd scores. Almost all hyperparameter combinations have all variables with an SMD score below 0.1, and the maximum SMD scores are very below (mostly between 0.04-0.08). This is significantly better than the SMD metrics of the RF models. However we shouldn't get too excited, since when looking at the number of matched pairs we quicky observe that the number of matched pairs is siginicantly lower (a lot between 900-1400) than in the RF models (1650-1700). This means that as much as half of the treated patients (2184 total) are thrown away during matching, which consequently means a smaller effective sample for the final outcome analysis (t-test, McNemar, logistic regression) , and thus wider confidence intervals, noisier estimates and less power to detect a real effect. Moreover, as explained before, matching estimates the ATT, the effect for the treated population. So, if a caliper throws away too many treated patients, then the estimate no longer represents the effect for all treated patients, but only the effect for the subset of treated patients who happened to be easy to match. That subset may not be representative, e.g. if the dropped patients are systematically different (they likely have more extreme propensity scores, meaning the least/most severely ill), then the ATT estimate is silently shifted toward a different, making the conclusions/finding less generalizable.


However, the SMD balancing metrics do seem to go hand-in-hand with the amount of overlap/positivity (see the graphs below for a representable subset of the propensity distributions across neural network settings) / the number of matched pairs (see tables above). In the graphs below we can clearly see a trend that the higher the lower the layer size (more so than decay magnitude), the more overlap there is (and thus consequently more matched pairs as seen in the tables above). Notably, the SMD results seem to follow a similar trend. In the heatmaps we can clearly see that the smaller models, as well as the more regularized models (larger weight decay) have better SMD metrics (all variables below 0.1 and smaller max SMD), and that in addition to the more matches we just noted. This can be explained by the fact that, as stated earlier, the dataset is very small and the prediction task at hand fairly simple, thus the model quickly becomes too powerfull, in the sense that it is able to predict the propensity scores too well (towards 0 and 1 for control/treated respectively, as can be very well seen in the graphs below for size = 12), which leads to very little overlap and thus very few & poor matches and thus bad balancing. A smaller model as well as a more regularized model, can predict and thus spread out the propensity scores less well, leading to more overlap and consequently better matching. The model has very similar (near optimal) SMD metrics for hidden sizes 1 and 2 (max smd of 0.041 vs 0.04), however, we do note in the second set of overlap/positivity graphs below that the model with hidden size = 1 has significantly better overlap which lead to approximately 100 more matches (approximatelty 1600-1650 vs. 1700-1750 as seen in the table above). Therefore, for our final model we choose the configuration of a hidden layer size of 1 and weight decay of 0.01 (most matches at 1743 AND best max smd score). 


```{r}
par(mfrow = c(3, 3))

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size2_decay0.1"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 2, decay = 0.1", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size2_decay0.001"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 2, decay = 0.001", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size2_decay0"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 2, decay = 0", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size6_decay0.1"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 6, decay = 0.1", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size6_decay0.001"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 6, decay = 0.100", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size6_decay0"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 6, decay = 0", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size12_decay0.1"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 12, decay = 0.1", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size12_decay0.001"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 12, decay = 0.001", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size12_decay0"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 12, decay = 0", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

par(mfrow = c(1, 1))


par(mfrow = c(2, 3))
pscore.nn <- pmin(pmax(ps.by.combo.nn[["size1_decay0.1"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 1, decay = 0.1", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size1_decay0.001"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 1, decay = 0.001", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size1_decay0"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 1, decay = 0", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size2_decay0.1"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 2, decay = 0.1", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size2_decay0.001"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 2, decay = 0.001", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

pscore.nn <- pmin(pmax(ps.by.combo.nn[["size2_decay0"]], 0.005), 0.995)
brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.nn, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.nn[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "size = 2, decay = 0", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.nn[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram

par(mfrow = c(1, 1))
```


Lastly, we get to the CBPS model. As explained before, the actual goal is to create well-balanced treated vs control groups. The regular glm does not directly optimize for this since it simply fits the model to maximize the likelihood of correctly predicting treatment. Balance is never part of the fitting objective and is only measured post-hoc by checking SMD. CBPS (Imai & Ratkovic, 2014) fix this by changing directly what the model is optimizing.

Instead of fitting the propensity model purely by maximum likelihood, the propensity score can be estimated in a way that maximizes a combination of covariate balance as well as the propensity prediction accuracy. CBPS uses the same parametric logistic form as ordinary logistic regression, but chooses the parameters by an augmented set of estimating equations that simultaneously satisfy the usual score/likelihood conditions from logistic regression, but also force the weighted covariate means to be equal between treated and control groups. In other words, balancing constraints are directly incorporated into the equations used to solve for the parameters, rather than only checked afterward. The method, therefore, avoids an iteration between model fitting and balance checking. Note that in the architecture of the CBPS model, we chose to target ATT (average treatment effect on the treated) rather than ATE since in general matching builds counterfactuals for only the treated group so all our estimates in this entire assignment have been estimating ATT and not ATE. Moreover, we use the "exact" method to ensure covariate means are matched exactly between groups.

```{r}
cbps.model <- CBPS(treatment.swang ~ ., data = rhc.selected[, c(xvars, "treatment.swang")],
                    ATT = 1, method = "exact") #Fit the CBPS model to the data
summary(cbps.model) #significance of coefficients etc. important for variable importance analysis later
pscore.cbps <- fitted(cbps.model)  # propensity scores
pscore.cbps.c <- pmin(pmax(pscore.cbps, 0.005), 0.995) #clipping as usual

brks <- seq(0, 1, 0.05)
ymax <- max(hist(pscore.cbps, breaks = brks, plot = FALSE)$counts)
#Plot propensity score frequencies for treated and untreated groups to check for overlap/positivity
hist(pscore.cbps[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
     xlab = "Propensity score", main = "Propensity score distribution in treated vs untreated groups.", ylim = c(0, ymax)) #Use transparansy of 0.6/0.5 to overlap/positivity them nicely
hist(pscore.cbps[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE) #Add to previous histogram to merge them into one overlap/positivityping histogram


logits.cbps <- logit(pscore.cbps.c)

set.seed(42) #seed for reproducibility if multiple controls are at same distance, as explained before.
# Same matching and SMD calculation steps as fore the general GLM logistic regression model earlier.
psmatch.cbps <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logits.cbps,
                       replace = FALSE, Weight = 2, caliper = 0.2)
matched_cbps <- rhc.selected[unlist(psmatch.cbps[c("index.treated", "index.control")]), ]
matchedtab1_cbps <- CreateTableOne(vars = xvars, strata = "treatment.swang", data = matched_cbps, test = FALSE)
print(matchedtab1_cbps, smd = TRUE)
smd_values_matched_cbps <- ExtractSmd(matchedtab1_cbps)

sprintf("Matched pairs: %d (of %d treated)", length(psmatch.cbps$index.treated), sum(rhc.selected$treatment.swang))
sprintf("Max SMD: %.3f", max(smd_values_matched_cbps))
sprintf("SMD < 0.1: %d variables", sum(smd_values_matched_cbps < 0.1))
sprintf("0.1 <= SMD < 0.2: %d variables", sum(smd_values_matched_cbps >= 0.1 & smd_values_matched_cbps < 0.2))
sprintf("SMD > 0.2: %d variables", sum(smd_values_matched_cbps > 0.2))

```
The results above show promising results. First of all, the overlap/positivity seems okay; both groups have their own slight peak but their is sufficient overlap present across a wide range from 0.1-0.9, which is confirmed by the relatively large number of matches at 1733. Moreover, the SMD balancing metrics are very positive, as all 40 variables scored below 0.1, and the max SMD even is 0.058. Overall, this means that this method, which bakes balancing directly into its optimization equations, indeed succeeds well at balancing the two groups across the covariates, while keeping a good sample size for the downstream effect analysis. 


# Now, we compare the number of matched pairs and balancing (through max SMD score, and number of variables below 0.1, between 0.1-0.2 and over 0.2 respecitvely) between each of the 5 methods. Below we report these metrics, as well as their propensity overlap/positivity graphs for the 4 propensity-based methods one more time for convenience.

Now that we have looked at all 5 methods seperately, and chose the best models for the two methods that required hyperparameter optimization (RF & NN). We can compare the balancing performance methods between the different methods. Below the overlap/positivity graphs of the 4 propensity-based methods are shown, as well as the number of matching pairs and SMD metrics for all 5 methods. The intial matching based on the Mahalanobis distance of the covariate directly used no caliper (as per instructions), therefore it matched all treated subjects (2184), which is a plus since it keeps the sample size as large as possible for the downstream final outcome analysis in the next part. However, to obtain comparable results, we performed greedy matching also with the 0.2 caliper (see below), which resulted in 0 matches. This means that not a single treated subject could find a control patient where each of the covariate values were at least 0.2 standardized units apart. This shows that yes it is evident that the greedy matching produces far less good matches compared to the other methods (as we will also see in the SMD metrics next). However, it might also suggest that the default 0.2 standardized units caliper might be a too strict metric in a high number of dimensions. It requires the mean of each of the 40 variables to be less than 0.2 standardized units away, possibly throwing away good potential matches that might be just too far away in one or a couple of the 40 dimensions. We will return to this and analyze it later when doing a caliper sensitivity analysis. Looking at the other methods in the table, we see much better number of matches. The GLM, NN and CPS all have a roughly equal number of matches (1730-1740), while the RF has a (slightly) lower number of matches (1659), of the total 2184 treated patients. When looking at the overlap graphs, we do see that the NN has a bit worse overlap than the rest, which can be explained by its enormous power and flexibility to produce very accurate propensity predicitions as we discussed before. However, overall the overlap/positivity assumption seems valid in all four methods. Lastly, when looking at the SMD results, we clearly observe that the greedy matching performs worst, with a max SMD of 0.501, only 24 variables with a score below 0.1, and 7 variables with a score above 0.2, indicating that the groups are highly inbalanced on those covariates. The other 4 methods score roughly similar, each having a max SMD score below 0.1 (implyinh that all 40 variables are below 0.1), with the original GLM method doing as low as 0.035. Based on these results, there does not seem to be a clear winner. However, given that they perform similarly, we prefer the original GLM implementation, as the more simplistic model is more interpretable, as we can more directly interpret it's coefficients to see which variables contributed in what way to the predictions, allowing for better explainability of the model. The original clearly greedy matching model clearly underperforms in all aspects to the other models, as discussed before.

```{r}
#Perform greedy matching also with the same caliper for comparable results
greedymatch2 <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = as.matrix(rhc.selected[xvars]), 
                      replace = FALSE, Weight = 2, caliper=0.2)
if (!is.list(greedymatch2) || is.null(greedymatch2$index.treated)) { #safety check to see if any matches at all
  print("No valid matches found with greedy matching with the 0.2 caliper.")
} else {
  sprintf("Greedy matching has %d matches when also using a 0.2 caliper", length(greedymatch2$index.treated))
}

# Pull the RF and NN SMD summary values from their comparison tables
rf.row <- rf.comparison[rf.comparison$nodesize == 10 & rf.comparison$mtry == 3, ]
nn.row <- nn.comparison[nn.comparison$size == 1 & nn.comparison$decay == 0.01, ]

sprintf("Matched pairs: Mahalanobis (no caliper) = %d, glm PS = %d, RF PS = %d, NN PS = %d, CBPS PS = %d (of %d treated)",
        length(greedymatch$index.treated), length(psmatch$index.treated),
        rf.row$matched.pairs,nn.row$matched.pairs, length(psmatch.cbps$index.treated), sum(rhc.selected$treatment.swang))

sprintf("Max SMD: Mahalanobis (no caliper) = %.3f, glm PS = %.3f, RF PS = %.3f, NN PS = %.3f, CBPS PS = %.3f",
        max(smd_values_matched), max(smd_values_matched_ps), rf.row$max.smd, nn.row$max.smd, max(smd_values_matched_cbps))

sprintf("SMD < 0.1: Mahalanobis (no caliper) = %d, glm PS = %d, RF PS = %d, NN PS = %d, CBPS PS = %d",
        sum(smd_values_matched < 0.1), sum(smd_values_matched_ps < 0.1), rf.row$n.smd.under.0.1, nn.row$n.smd.under.0.1, sum(smd_values_matched_cbps < 0.1))

sprintf("0.1 <= SMD <= 0.2: Mahalanobis  (no caliper)= %d, glm PS = %d, RF PS = %d, NN PS = %d, CBPS PS = %d",
        sum(smd_values_matched >= 0.1 & smd_values_matched <= 0.2), sum(smd_values_matched_ps >= 0.1 & smd_values_matched_ps <= 0.2), rf.row$n.smd.0.1.0.2, nn.row$n.smd.0.1.0.2, sum(smd_values_matched_cbps >= 0.1 & smd_values_matched_cbps <= 0.2))

sprintf("SMD > 0.2: Mahalanobis (no caliper) = %d, glm PS = %d, RF PS = %d, NN PS = %d, CBPS PS = %d",
        sum(smd_values_matched > 0.2), sum(smd_values_matched_ps > 0.2),
        rf.row$n.smd.over.0.2, nn.row$n.smd.over.0.2, sum(smd_values_matched_cbps > 0.2))

pscore.rf <- pmin(pmax(ps.by.combo[["ns10_mtry3"]], 0.005), 0.995)
pscore.nn <- pmin(pmax(ps.by.combo.nn[["size1_decay0.01"]], 0.005), 0.995)

par(mfrow = c(2, 2))
pscores.list <- list(
  "GLM"  = pscore,
  "RF"   = pscore.rf,
  "NN"   = pscore.nn,
  "CBPS" = pscore.cbps
)

brks <- seq(0, 1, 0.05)

for (name in names(pscores.list)) {
  ps <- pscores.list[[name]]
  ymax <- max(hist(ps, breaks = brks, plot = FALSE)$counts)
  
  hist(ps[rhc.selected$treatment.swang == 1], breaks = brks, col = rgb(0, 0, 0.55, 0.6),
       xlab = "Propensity score", main = paste(name, "propensity score overlap/positivity"), ylim = c(0, ymax))
  hist(ps[rhc.selected$treatment.swang == 0], breaks = brks, col = rgb(0.55, 0, 0, 0.5), add = TRUE)
}

par(mfrow = c(1, 1))
```


# Here we define our best performing model for each of the 2 finetuned methods and get their corresponding matched sets, so we can then continue with the rest of the analysis to see whether treatment significantly explains outcomes for each of the methods.
```{r}
# Create matched set for final random forest (node size = 10, mtry = 3 , 300 trees) for further downstream analysis.
set.seed(42)
# Final, best balancing Random Forest:
rf.model  <- randomForest(treatment.swang ~ ., data = rf.data, ntree = 300, nodesize = 10, mtry = 3, importance = TRUE)
pscore.rf <- predict(rf.model, type = "prob")[, "1"]
psmatch.rf <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logit(pmin(pmax(pscore.rf, 0.005), 0.995)),
                    replace = FALSE, Weight = 2, caliper = 0.2)
matched_rf <- rhc.selected[unlist(psmatch.rf[c("index.treated", "index.control")]), ]


# Create matched set for final neural network (weight decay = 0.01, hidden size = 1) for further downstream analysis.
# Final, best balancing Neural Network:
m <- nnet(treatment.swang ~ ., data = nn.df, size = 1, decay = 0.01,
          trace = FALSE, maxit = 300, linout = FALSE) 
pscore.nn <- predict(m, newdata = nn.df, type = "raw")[, 1] 
psmatch.nn <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = logit(pmin(pmax(pscore.nn, 0.005), 0.995)), replace = FALSE, Weight = 2, caliper = 0.2)
matched_nn <- rhc.selected[unlist(psmatch.nn[c("index.treated", "index.control")]), ]
```



#Next, we perform the downstream outcome statistical analysis for each of the three new methods. Note that even though we prefer the original GLM model for its simplicity and interpretibility, we perform the analysis for all three models for comparison anyways, since the balancing performance was very similar.

```{r}
print("===================================================")
set.seed(42) #For reproducibility of results
# First for the RF:
print("For the RF:")
y_trt.rf <- matched_rf$outcome.died[matched_rf$treatment.swang == 1]
y_con.rf <- matched_rf$outcome.died[matched_rf$treatment.swang == 0]

t.test(y_trt.rf - y_con.rf)                 # paired t-test, indicative only
table(y_trt.rf, y_con.rf)
mcnemar.test(table(y_trt.rf, y_con.rf))     # correct test for paired binary data

gmodel5 <- glm(outcome.died ~ treatment.swang, family = binomial, data = matched_rf)
summary(gmodel5)

# Doubly robust: adjust for the least well balanced variable
worst.rf <- rf.row$worst.var
gmodel6 <- glm(as.formula(paste("outcome.died ~ treatment.swang +", worst.rf)), family = binomial, data = matched_rf)
summary(gmodel6)
sprintf("Odds ratio for treatment: %.4f", exp(coef(gmodel5)["treatment.swang"]))
sprintf("Odds ratio for treatment after adjusting for %s: %.4f", worst.rf, exp(coef(gmodel6)["treatment.swang"]))

effects <- data.frame(
  method      = c("Mahalanobis", "PS (glm)", "PS (random forest)"),
  pairs       = c(length(greedymatch$index.treated), length(psmatch$index.treated), length(psmatch.rf$index.treated)),
  mean.diff   = c(mean(y_trt - y_con), mean(y_trt.ps - y_con.ps), mean(y_trt.rf - y_con.rf)),
  OR          = exp(c(coef(gmodel1)["treatment.swang"], coef(gmodel3)["treatment.swang"], coef(gmodel5)["treatment.swang"])),
  OR.adjusted = exp(c(coef(gmodel2)["treatment.swang"], coef(gmodel4)["treatment.swang"], coef(gmodel6)["treatment.swang"])))
#print(effects, digits = 4, row.names = FALSE)


# Next, for the neural network:
print("For the NN:")
y_trt.nn <- matched_nn$outcome.died[matched_nn$treatment.swang == 1]
y_con.nn <- matched_nn$outcome.died[matched_nn$treatment.swang == 0]

t.test(y_trt.nn - y_con.nn)
mcnemar.test(table(y_trt.nn, y_con.nn))

gmodel7 <- glm(outcome.died ~ treatment.swang, family = binomial, data = matched_nn)
summary(gmodel7)
worst.nn <- nn.row$worst.var
gmodel8 <- glm(as.formula(paste("outcome.died ~ treatment.swang +", worst.nn)), family = binomial, data = matched_nn)
summary(gmodel8)

sprintf("OR (unadjusted): %.4f", exp(coef(gmodel7)["treatment.swang"]))
sprintf("OR (adjusted for %s): %.4f", worst.nn, exp(coef(gmodel8)["treatment.swang"]))

effects <- rbind(effects, data.frame(
  method       = "PS (neural net)",
  pairs        = length(psmatch.nn$index.treated),
  mean.diff    = mean(y_trt.nn - y_con.nn),
  OR           = exp(coef(gmodel7)["treatment.swang"]),
  OR.adjusted  = exp(coef(gmodel8)["treatment.swang"])))
#print(effects, digits = 4, row.names = FALSE)

#Lastly, for the CPBS model:
print("For the CBPS:")
y_trt.cbps <- matched_cbps$outcome.died[matched_cbps$treatment.swang == 1]
y_con.cbps <- matched_cbps$outcome.died[matched_cbps$treatment.swang == 0]

t.test(y_trt.cbps - y_con.cbps)
mcnemar.test(table(y_trt.cbps, y_con.cbps))

gmodel9 <- glm(outcome.died ~ treatment.swang, family = binomial, data = matched_cbps)
summary(gmodel9)
worst.cbps <- rownames(smd_values_matched_cbps)[which.max(smd_values_matched_cbps)]
gmodel10 <- glm(as.formula(paste("outcome.died ~ treatment.swang +", worst.cbps)), family = binomial, data = matched_cbps)
summary(gmodel10)

sprintf("OR (unadjusted): %.4f", exp(coef(gmodel9)["treatment.swang"]))
sprintf("OR (adjusted for %s): %.4f", worst.cbps, exp(coef(gmodel10)["treatment.swang"]))

effects <- rbind(effects, data.frame(
  method       = "CBPS",
  pairs        = length(psmatch.cbps$index.treated),
  mean.diff    = mean(y_trt.cbps - y_con.cbps),
  OR           = exp(coef(gmodel9)["treatment.swang"]),
  OR.adjusted  = exp(coef(gmodel10)["treatment.swang"])))
print(effects, digits = 4, row.names = FALSE)


print("===================================================")
```
For the RF model, the p-value values are all significant: for the t-test 0.0411, the McNemar test 0.04489, and the coefficient of the glm model has a significance value of 0.0398 and results in an odds ratio for treatment: 1.1632. After explicitly adjust for the highest SMD variable (Cirrhosis), the coefficient p-value actually decreased to 0.02805, and the odds ratio rises to 1.1757. For the NN model, the effect seems to be even stronger in all tests: the p-value for the t-test is 0.000416, for the McNemar test 0.000483, and the coefficient of the glm model has a significance value of 0.000324 and results in an odds ratio for treatment: 1.2929. After explicitly adjust for the highest SMD variable (wtkilo1), the coefficient p-value actually decreased to 0.000206, and the odds ratio again rises slightly to 1.3047. Lastly, for the CBPS model the treatment effects become even stronger: the p-value for the t-test is 4.184e-05, for the McNemar test 5.048e-05, and the coefficient of the glm model has a significance value of 4.36e-05  and results in an odds ratio for treatment: 1.3390. After explicitly adjust for the highest SMD variable (wtkilo1), the coefficient p-value is 8.35e-05, and the odds ratio again rises slightly to 1.3260. Comparing this to the original GLM propensity model, we see that the NN and CBPS obtain similar effect results, and also obtain similar mean difference results (close to the glm's approximate 0.06). The RF shows the mean difference lower at 0.03. However, all models still agree that there is significant evidence that treatment does affect outcome, and in fact the chance of dying increases with RHC treatment. Since the 4 propensity-based models balanced the data similarly well, we place roughly equal amounts of trust in their results. However, since we chose the original GLM-based propensity method as our preferred, conditional on our selected covariates capturing the relevant confounding, the glm-propensity-matched result is our best estimate, because it achieved the (marginally) best-balanced comparison among the methods we tested. Notably, the forest did not confirm our hypothesis that a threshold-based model would balance better, but it and all the other models do earn their place as a robustness check: they all still find higher mortality among treated patients, so the RHC result does not depend on the form of the logistic model Mahalanobis matching is the clear loser, with the most group imbalance and the estimate that shifted most under adjustment.


# Next, since the 4 propensity models balance similarly well, we want to try and dive a little bit deeper into why this is the case. Do they balance similarly well because they converge on the same distribution of propensity scores? Or do they each get different propensity scores but do the different scores happen to balance simiarly well. As stated at the start, we have no real interest in the propensity scores themselves, only their usage as a balancing tool, so probing this can help us understand the balancing mechanism better. 
```{r}
pscores.df <- data.frame(
  glm  = pscore,
  rf   = pscore.rf,
  nn   = pscore.nn,
  cbps = pscore.cbps
)
cor.matrix.pearson <- cor(pscores.df) #Perform Pearson (linear) correlation
cor.matrix.spearman <- cor(pscores.df, method = "spearman") #Perform Spearman (rank) correlation to check for relative ordering and not just numeric scale.

print(cor.matrix.pearson)
print(cor.matrix.spearman)

corrplot(cor.matrix.pearson, method = "color", type = "upper", addCoef.col = "black",
         title = "Pearson correlation between propensity scores across methods", mar = c(0,0,1,0))

corrplot(cor.matrix.spearman, method = "color", type = "upper", addCoef.col = "black",
         title = "Spearman correlation between propensity scores across methods", mar = c(0,0,1,0))

```
These correlation results reveal a clear pattern: the glm, NN, and CBPS propensity scores are all very highly correlated with each other (Pearson/Spearman both approximately 0.98–0.99), while RF stands notably apart, correlating with the other three at only 0.83–0.84. Note that both Pearson and Spearman agree with each other everywhere, so this isn't a scale-vs-ranking artifact, rather the RF genuinely ranks patients differently compared to the other models, not just on a different numeric scale. Importantly, this is consistent with our earlier reasoning about the models their functional forms: glm, NN, and CBPS all model treatment as some smooth, additive function of the covariates, while RF's completely different threshold-based, interaction-capturing splits produce a very different picture of who's likely treated. This divergence here is in correspondance with RF's comparatively worse SMD performance throughout our analysis; the model that thinks differently about the data is also the one that balances covariates least well, suggesting its threshold-based reasoning, whatever its clinical plausibility, isn't capturing the confounding structure as effectively as the smoother models are.



## Variable importance

To further probe this interesting difference in propensity value prediction strategies, we want to find out why and how each approach determines who gets treated. To do this, we will compare the variable importance of the GLM (our preferred model) and RF propensity prediction models.

For the glm we cannot simply compare the raw coefficients, because a coefficient is per unit of its variable (one year of age versus one point of APACHE score versus a 0/1 indicator), so instead we use the absolute z-value (coefficient divided by its standard error), which is scale-free and is exactly what the p-value in summary(psmodel) is based on. For the forest we use both the mean decrease in Gini impurity, i.e. how much splitting on a variable improved the purity of the treated/untreated mix, summed over every split in every tree, and the permutation importance which randomly permutes the values of a variables (removing any statisticaly relationship with the outcome) and then computes the decrease in accuracy. We do this because one caveat has to be kept in mind is that Gini importance is biased towards continuous variables, because a variable like meanbp1 or age offers a tree hundreds of candidate cutoffs whereas a 0/1 indicator like dnr1 offers exactly one, so the continuous variable has more possibilities to create an optimal split (and therefore be important), so the forest will naturally rank the continuous physiological measurements above the binary history and category variables. Lastly, since the two measures are on completely different scales, so we compare ranks rather than values.

```{r}
print("===================================================")
set.seed(42)
glm.abs.z <- abs(summary(psmodel)$coefficients[-1, "z value"])  # [-1] drops the intercept
imp.all   <- importance(rf.model)
rf.gini   <- imp.all[, "MeanDecreaseGini"]
rf.perm   <- imp.all[, "MeanDecreaseAccuracy"]

importance.comparison <- data.frame(
  glm.abs.z = round(glm.abs.z[xvars], 2),
  rf.gini   = round(rf.gini[xvars], 1),
  rf.perm   = round(rf.perm[xvars], 2)
)

importance.comparison$glm.rank      <- rank(-importance.comparison$glm.abs.z)  # rank 1 = most important
importance.comparison$rf.gini.rank  <- rank(-importance.comparison$rf.gini)
importance.comparison$rf.perm.rank  <- rank(-importance.comparison$rf.perm)

print(importance.comparison[order(importance.comparison$rf.gini.rank), ])

sprintf("Spearman correlation, glm vs RF Gini: %.2f",
        cor(importance.comparison$glm.abs.z, importance.comparison$rf.gini, method = "spearman"))
sprintf("Spearman correlation, glm vs RF permutation: %.2f",
        cor(importance.comparison$glm.abs.z, importance.comparison$rf.perm, method = "spearman"))
sprintf("Spearman correlation, RF Gini vs RF permutation: %.2f",
        cor(importance.comparison$rf.gini, importance.comparison$rf.perm, method = "spearman"))

varImpPlot(rf.model, n.var = 15, type = 1,
           main = "Random forest permutation importance (top 15)", cex = 0.6)
varImpPlot(rf.model, n.var = 15, type = 2,
           main = "Random forest Gini importance (top 15)", cex = 0.6)

print("===================================================")
```
The rank correlation tells us to what degree the two models look at the same variables: a value near 1 would mean they weigh the covariates in the same order as predictor, a value near 0 that their is no monotonic relationship between the ordering of the importance of the variables: knowing how important a variable is to glm gives you no information about how important it is to RF (and vice versa). A variable that the forest ranks high but the glm ranks low, and that is not simply continuous, is therefore the best candidate for a threshold or interaction effect that the additive glm cannot express. Since the balance after matching was nonetheless better for the GLM, whatever extra structure the forest picked up did not translate into a better propensity score, which is the same conclusion as before seen from the other side. 
In the results we can actually see a very interesting pattern. There are a few stand out disagreements (e.g. rea1 (creatinine): glm rank 30 (barely significant, z=1.11) vs. RF rank 3 (one of its top variables)) and a lot of moderate disagreements (e.g age: glm rank 38 (z=0.44, essentially not significant) vs. RF rank 6), resulting in a Spearman ranking correlation of only 0.14 for Gini and 0.33 for correlation. It initially sounds quite surprising that there is practically no correlation between the variable importance ranks, even though the propensity scores correlate at the much more significant 0.83. However, when looking closer we can see that the models agree on a small set of dominant predictors. For example, meanbp1 and aps1 both being in the top 1-3, having large z of approx. 8). These few dominant predictor dominate the predictions, which is enough to drive fairly high score correlation. Why certain variables are important for the RF model but not for the GLM (e.g. crea1 and age), can be because those variables do not correlate linearly with treatment, but once included in non-linar interactions or splits/thresholds, they can indeed become important. Lastly, it is notable that their is only a Spearman correlation of 0.61 between the RF Gini ranking and RF permutation ranking, indicating that the Gini importance was partly inflating the apparent RF/glm disagreement, likely due to its known bias toward continuous variables with many split points (as explained earlier). Once you switch to the less-biased permutation measure, RF and glm's importance rankings converge somewhat, though still far from strong agreement (0.33 is still weak).

#Partial dependence plot
To investigate why RF and glm disagreed sharply on dominant predictors for RF like crea1 and age (high RF importance, near-zero glm significance), we plotted partial dependence curves for both variables to try and probe/interpret the RF's model structure. This holds all other covariates at their observed values while sweeping the variable in questions across a range of values, and averaging each model's predicted propensity score. Both variables show visible nonlinear and threshold/step behavior in the RF, not the smooth trend a linear glm term would produce — confirming RF is genuinely capturing nonlinear structure the glm cannot express.
This is a nice nuance on our earlier finding that RF balanced worse than GLM despite this hypothesis. RF's threshold behavior is real, it just didn't help matching. This is likely because it also produced a rougher, more separated score distribution (consistent with RF's higher extreme-score counts and less correlated scores we saw before). So the intial "RF should balance better because it thinks like a clinician" hypothesis was right about the mechanism, but wrong about the practical payoff, which is a more interesting result, since it separates whether a model captures the expected structure from whether that structure helps the specific task of matching.

```{r}
par(mfrow = c(1, 2))

partialPlot(rf.model, pred.data = rf.data, x.var = "crea1",
            main = "Partial dependence: crea1", xlab = "Creatinine (crea1)", ylab = "logit(P(treated))")

partialPlot(rf.model, pred.data = rf.data, x.var = "age",
            main = "Partial dependence: age", xlab = "Age", ylab = "logit(P(treated))")

par(mfrow = c(1, 1))
```


## Caliper width

Throughout the assignment we matched with a caliper of 0.2 standard deviations of the logit of the propensity score, which is the width recommended by Austin (2011), but we never checked whether that choice mattered, it was simply given to us by the instructors. The caliper is the knob that sets the bias-variance tradeoff of matching directly: a narrow caliper only accepts close matches, so the matched groups are well balanced (low bias), but more treated patients are left without a match, so the outcome effect estimate is based on fewer pairs (high variance) and, more subtly, on a different group of patients than the full treated population, as explained before. On the other hand, a wide caliper keeps nearly all treated patients but accepts poorer matches, so the imbalance creeps back in. Below we repeat the matching for a range of calipers, for all four propensity score models, and record the number of pairs, the balance and the resulting effect estimate given as the odds ratio from the GLM coefficient. Later, we perform a similar analysis for the greedy matching approach, but we do this seperately since it lives on an entirely different caliper scale as it is in 40D rather than 1D. 

```{r}
calipers <- c(0.05, 0.1, 0.2, 0.5, 0.75, 1, 2)
logits.rf   <- logit(pmin(pmax(pscore.rf, 0.005), 0.995))
logits.nn   <- logit(pmin(pmax(pscore.nn, 0.005), 0.995))
logits.cbps <- logit(pmin(pmax(pscore.cbps, 0.005), 0.995))

caliper.comparison <- data.frame()

for (cal in calipers) {
  for (model in c("glm", "rf", "nn", "cbps")) {
    X.model <- switch(model,
                       glm  = logits,
                       rf   = logits.rf,
                       nn   = logits.nn,
                       cbps = logits.cbps)
    set.seed(42)
    mt <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = X.model,
                replace = FALSE, Weight = 2, caliper = cal)
    md  <- rhc.selected[unlist(mt[c("index.treated", "index.control")]), ]
    smd <- ExtractSmd(CreateTableOne(vars = xvars, strata = "treatment.swang", data = md, test = FALSE))
    y.t <- md$outcome.died[md$treatment.swang == 1]
    y.c <- md$outcome.died[md$treatment.swang == 0]
    
    gmodel.cal <- glm(outcome.died ~ treatment.swang, family = binomial, data = md)
    coef.summary <- summary(gmodel.cal)$coefficients

    caliper.comparison <- rbind(caliper.comparison, data.frame(
      model          = model,
      caliper        = cal,
      matched.pairs  = length(mt$index.treated),
      max.smd        = round(max(smd), 3),
      n.smd.over.0.1 = sum(smd > 0.1),
      mean.diff      = round(mean(y.t - y.c), 3),
      OR             = round(exp(coef.summary["treatment.swang", "Estimate"]), 3),
      p.value        = signif(coef.summary["treatment.swang", "Pr(>|z|)"], 3)))
  }
}
print(caliper.comparison, row.names = FALSE)

ggplot(caliper.comparison, aes(x = caliper, y = matched.pairs, colour = model)) +
  geom_line() + geom_point() + scale_x_log10() +
  labs(x = "caliper (SD of logit PS, log scale)", y = "matched pairs", title = "Matched pairs kept vs caliper width") +
  theme_minimal()

ggplot(caliper.comparison, aes(x = caliper, y = max.smd, colour = model)) +
  geom_line() + geom_point() + scale_x_log10() +
  geom_hline(yintercept = 0.1, linetype = "dashed") +
  labs(x = "caliper (SD of logit PS, log scale)", y = "max SMD", title = "Worst-variable SMD vs caliper width") +
  theme_minimal()
```
The two plots are the bias-variance tradeoff drawn out: the number of pairs rises with the caliper while the maximum SMD rises with it too, and the question is where the curves cross the point of diminishing returns. The sensible choice is the widest caliper that still keeps every variable under the dashed 0.1 line, since beyond that we are getting pairs with significant imbalance, and before that we are throwing away pairs without gaining much balance as the max-SMD plot is essentially flat for the first half. Notably, we see that the random forest model has a worse max SMD score and drops pairs faster throughout, highlighting its worse balancing performance as we demonstrated earlier. For the other three models, the optimal caliper seems to be around 0.75, which actually would have increased the number of matched pairs by around 200 for those models, significantly increasing the sample size to allow for smaller confidence intervals, less noisy estimates and more power to detect a real effect. Notably, in the results table we see that at this larger sample size the p-values are all still significant (5.65e-04 vs. 2.81e-02 vs. 1.16e-03 vs. 2.45e-05) and the odds ratios still indicate that the chance of dying increases with treatment (OR 1.266 vs. 1.166 vs. 1.247 vs. 1.334) for the GLM vs. RF vs. NN vs. CBPS. respectively.


Regarding the greedy matching with the Mahalanobis distance; we already observed that it obtains 0 matches using a caliper due to the matching happening in 40D. Moreover, it is interpreted as the variation in each covariate in standardized units of that covariate, rather than the variation in the logit of a predicted scalar - the caliper means very different things in each of the cases. As a result, the caliper range needsa completely different variation scale and shouldn't be compared directly with the other calipers, which is why we gave it it's own brief analysis here. 
```{r}
calipers.maha <- c(0.5, 1, 2, 5, 10, 20)  # far larger, since caliper applies per-covariate across 40 dimensions
maha.comparison <- data.frame()

for (cal in calipers.maha) {
  set.seed(42)
  mt <- Match(Tr = rhc.selected$treatment.swang, M = 1, X = as.matrix(rhc.selected[xvars]),
              replace = FALSE, Weight = 2, caliper = cal)
  
  if (!is.list(mt) || is.null(mt$index.treated)) {
    maha.comparison <- rbind(maha.comparison, data.frame(
      caliper = cal, matched.pairs = 0, max.smd = NA, n.smd.over.0.1 = NA, mean.diff = NA, OR = NA))
    next
  }
  
  md  <- rhc.selected[unlist(mt[c("index.treated", "index.control")]), ]
  smd <- ExtractSmd(CreateTableOne(vars = xvars, strata = "treatment.swang", data = md, test = FALSE))
  y.t <- md$outcome.died[md$treatment.swang == 1]
  y.c <- md$outcome.died[md$treatment.swang == 0]
  
  gmodel.cal <- glm(outcome.died ~ treatment.swang, family = binomial, data = md)
  coef.summary <- summary(gmodel.cal)$coefficients

  maha.comparison <- rbind(maha.comparison, data.frame(
    caliper        = cal,
    matched.pairs  = length(mt$index.treated),
    max.smd        = round(max(smd), 3),
    n.smd.over.0.1 = sum(smd > 0.1),
    mean.diff      = round(mean(y.t - y.c), 3),
    OR             = round(exp(coef.summary["treatment.swang", "Estimate"]), 3),
    p.value        = signif(coef.summary["treatment.swang", "Pr(>|z|)"], 3)))
}
print(maha.comparison, row.names = FALSE)
```
In the results table above, we can see a few interesting results. First, we see that as the caliper is increased significantly, in range of 5+, we do get a significant amount of matches, meaning that in that case most treated patients do find at least 1 (unmatched) control where each of the 40 variables differ by no more than 5 standardized units. Notably, the p-values here also get small (in the range of e-10), and the odds ratio indicates, again, a positive result (death chances increase with treatment), though the OR is a lot larger than before: approximately 1.5 here whereas the other model were in the range of 1.1-1.3. However, we also observe that the number of variables with SMD above 0.1 is still 16 (which is more than at smaller calipers since the matches allowed are worse, per the bias-variance tradeoff discussed above), and the max smd stays around 0.5, which raises significant concerns regarding confounding. Considering this, even at higher calipers, getting larger amounts of matches, we still prefer the other models as the balancing of the covariates between the two groups is significantly better for those methods, allowing us to place significantly more trust in our obtained results.

# Inverse probability weighting (IPW)
Throughout this entire report, we simply assumed the matching strategy given to us at the start. However, maybe this is not the optimal strategy to use and do other strategies achieve better or similar balancing quality, while preserving more matched pairs. As such, as a last experiment, we try out an alternative to matching: inverse probability weighting. Here, all treated observations get a weight of 1, while the untreated observations get weighted by their odds of being treated (odds of the propensity score). Thus, untreated patients who look very likely to be treated are upweighted, since they are likely rare among the untreated, while untreated patients which don't look like treated pateints get downweighted, since they are common among this group. As a result, the weighted distribution of covariates of the untreated group is reshaped to look very similar to that of the treated group, (hopefully) achieving balancing of each covariates, the same result that matching tries to achieve. Then, all patients are used in a weighted manner in the downstream outcome effect analysis. This allows us to isolate and compare each model's own balancing quality, without relying on the matching step to drop bad quality matches. 
```{r}
compute_att_weights <- function(ps, treatment) {
  ifelse(treatment == 1, 1, ps / (1 - ps))
}

pscores.list <- list(
  glm  = pscore,
  rf   = pscore.rf,
  nn   = pscore.nn,
  cbps = pscore.cbps
)

ipw.comparison <- data.frame()

for (name in names(pscores.list)) {
  ps.c <- pmin(pmax(pscores.list[[name]], 0.005), 0.995)
  w <- compute_att_weights(ps.c, rhc.selected$treatment.swang)
  
  design.w <- svydesign(ids = ~1, data = rhc.selected[, c(xvars, "treatment.swang", "outcome.died")], weights = ~w)
  
  # Balance check as before
  tab.weighted <- svyCreateTableOne(vars = xvars, strata = "treatment.swang", data = design.w, test = FALSE)
  smd.weighted <- ExtractSmd(tab.weighted)
  
  # Test treatment effect on outcome, using the same weights design
  m <- svyglm(outcome.died ~ treatment.swang, family = quasibinomial, design = design.w)
  cf <- summary(m)$coefficients
  
  ipw.comparison <- rbind(ipw.comparison, data.frame(
    model           = name,
    max.smd         = round(max(smd.weighted), 3),
    n.smd.under.0.1 = sum(smd.weighted < 0.1),
    n.smd.0.1.0.2   = sum(smd.weighted >= 0.1 & smd.weighted < 0.2),
    n.smd.over.0.2  = sum(smd.weighted > 0.2),
    max.weight      = round(max(w), 1),
    OR              = round(exp(cf["treatment.swang", "Estimate"]), 3),
    p.value         = signif(cf["treatment.swang", "Pr(>|t|)"], 3)))
}
print(ipw.comparison, row.names = FALSE)
```
In these results we observe similar trends as before. Notably, the balancing of RF is, again, the worst with a max smd = 0.11 and 2 variables over 0.1. The other methods perform similar, however the glm propensity prediction model does seem to fall behind a bit with a max smd = 0.1 and 1 variables over 0.1, whereas with regular matching its max smd was only 0.035 (and thus 0 variables over 0.1). This might suggest that that model was helped relatively a lot by the matching step disposing of bad untreated patients. Moreover, here the p-values are also all significant and comparable to the ones obtained with matching, the same for the odds ratios. 
However, as a general approach matching does seem to be preferable over IPW, since it does lead to better matching and does not reduce the power of the outcome analysis significantly, as we just compared. 


# Final Conclusions
In conclusion, after deep and thoughtful analysis of different propensity prediction models used for matching, including probing of various properties of them, and looking at an alternatives to the standard matching architecture we were given initially, we arrive at one most favourable approach. The intial GLM (logistic regression) propensity prediction model is the preferred model of choice as it: 1. produced the best overall balancing of each covariate across the treated and non-treated groups, and 2. it is the more simplistic and consequently interpretable model, as we can look at its standardized coefficients to directly get an accurate idea of the role each variable plays in its predictions, as opposed to, for example, the neural network whose weights are much more a blackbox. Moreover, we observed that matching with a slightly better caliper (approximately 0.75 standardized units) would allow for more matches, and consequently more power and less noisy estimates in the downstream outcome analysis, while hardly reducing the balancing is the best balancing approach, and preferred over direct inverse probability weighting. This preferred approach led to an odds ratio of 1.266 with a p-value of 5.65e-04, suggesting that their is enough evidence to conclude that RHC treatment increase the odds of dying, assuming the ignorability assumption holds.



#References
Austin, P. C. (2011). Optimal caliper widths for propensity-score matching when estimating differences in means and differences in proportions in observational studies. Pharmaceutical Statistics, 10(2), 150-161.

Imai, K., & Ratkovic, M. (2014). Covariate balancing propensity score. Journal of the Royal Statistical Society Series B: Statistical Methodology, 76(1), 243-263.
